---
title: "Employment Rate"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

The employment rate is the percentage of people age 16 and over who are working. The data comes from U.S. Census Bureau, American Fact Finder table S2301. 

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

df <- read_csv('data/employment_all_years_counties.csv') %>%
  # we only need rows representing the employment rate
  # these rows have the phrase 'Employed;' or 'Employment/Population Ratio' in the description
  # change the description so all rows have phrase 'Employment/Population Ratio'; which is the most recent phrase
  mutate(description = str_replace(.$description, 'Employed; ', 'Employment/Population Ratio; ')) %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # only keep employment rate rows and total population counts
         # totals are needed to derive percentages from artificially created groups (age bins)
         str_detect(description, 'Employment[/]Population Ratio;|Total')) %>%
  # the last section of the variable name (label) column is the same for estiamtes and totals
  # this section needs to be extracted so that estimate and total data sets can be merged on it
  mutate(var_name = str_extract(label, 'VC[0-9][0-9]$'))

# make the total counts an additional column

# create new dataset with just totals
emp_total <- df %>%
  # filter for only totals
  filter(str_detect(description, 'Total'),
         # remove 'PERCENT IMPUTED' rows because these do not reflect counts
         !str_detect(description, 'PERCENT IMPUTED')) %>%
  # only keep the variables that are needed to merge this dataset with the primary dataset
  select(var_name, geo_description, estimate, moe, year)

# remove totals from primary data set and merge in newly created totals data set
df <- df %>%
  filter(!str_detect(description, 'Total')) %>%
  left_join(emp_total, by = c('var_name', 'geo_description', 'year')) %>%
  rename(estimate = estimate.x, moe = moe.x,
         population = estimate.y, population_moe = moe.y)
         
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall employment rate of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over') %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Employment rate',
         color = 'Comparison Unit',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Interpretation

* The Forsyth County's employment rate lags behind Durham County's and Guilford County's rates. But, the difference between Forsyth County and Guilford County is not statistically significant.
* Forsyth County's employment rate has increased the past three years, but this increase is not statistically significant.

#### Data

Below is the raw data for employment rates by geographies.

```{r include = to_include}
total_data <- df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over') %>%
  select(geo_description, year, estimate, moe, se, cv)

# this represents the final dataset that will be sent to others for visualizations
tableau_data <- total_data %>%
    # add type of column
    mutate(type = 'Employment Rate',
           subtype = 'Total Population')

# create table
ff_data_dt(total_data, c('Geographic area', 'Year', '% Employed', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))

```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Employment rate by ethnicity

#### Plot

The plots below show employment rate by ethnicity and year. Many Hispanic values are missing due to insufficient data.

```{r}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description)

# line plot of ethnicity with Forsyth and all years
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    #facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth County employment rate by ethnicity',
         color = 'Ethnicity',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=ethnicity)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Employment rate by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Employment Rate',
       x = '') +
  theme_minimal()
```

#### Interpretation

* For most years, there is insufficient data to analyze the employment rate for Hispanic/Latinx residents in Forsyth County.
* In Forsyth County, the 2017 difference between the White, non-Hispanic and African American rates is not statistically significant, but the 2016 difference is.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
ethnicity_data <- ethnicity  %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv)

# add ethnicity information t ofinal datasets
tableau_data <- ethnicity_data %>%
    # add type of column
    mutate(type = 'Race and Ethnicity',
           subtype = ethnicity) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-ethnicity) %>%
    bind_rows(tableau_data, .)

ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', '% Employed', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'ethnicity'))
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'ethnicity'))
```

------

### Employment rate by age

#### Plot

The following plots examines the employment rate by age group.

```{r}
# some age bins are artificially created; we will need to derive percentages for these bins

# creat bins for age
age <- df %>%
  filter(# only keep age
         str_detect(description, ' AGE - ')) %>%
  # create new variable that is only the age
  mutate(age = str_extract(description, 'AGE - .* years'),
         age = str_extract(age, '[0-9].*'),
         # create variable that is the number of people employed in each age bin
         # this is needed to calculate the percentage of people employed in the artificially created age bins
         # must divide estimate by 100 because it is a percentage, but listed as integers
         pop_employed = population * (estimate/100),
         # create function for calculating the moe of two products; will be used below
         # reference: A compass for understaing and using american community survey data, October 2008, A-16
         pop_employed_moe = sqrt((estimate/100)^2 * population_moe^2 + population^2 * (moe/100)^2)) %>%
  # reclassify ages based on new categories
  # mutate(age  = ifelse(.$age == "16 to 19 years" | .$age == "20 to 24 years" | .$age == "25 to 29 years", '16 to 29 years',
                       #ifelse(.$age == "30 to 34 years" | .$age == "35 to 44 years" |
                                #.$age == "25 to 44 years", '30 to 44 years', '45 years and over')))
    mutate(age = ifelse(.$age == "16 to 19 years" | .$age == "20 to 24 years" , '16 to 24 years',
                       ifelse(.$age == "25 to 29 years" | .$age == "30 to 34 years" | .$age == "35 to 44 years" |
                                .$age == "25 to 44 years", '25 to 44 years', 
                       ifelse(.$age == "45 to 54 years" | .$age == "55 to 64 years" | .$age == "55 to 64 years" |
                                .$age == "60 to 64 years" | .$age == "55 to 59 years", '45 to 64 years','65 years and older'))))


# The following section creates the employment rate and MOE of each artificially derived age bin

age <- age %>%
  # group by year, geography, and the newly created age bins
  group_by(year, geo_description, age) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(population),
            sum_pop_emp = sum(pop_employed),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(population_moe, population),
            sum_pop_emp_moe = moe_sum(pop_employed_moe, pop_employed)) %>% 
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_employed = round( 100 * (sum_pop_emp / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_employed_moe = moe_prop(sum_pop_emp, sum_pop, sum_pop_emp_moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_employed_moe = round( perc_employed_moe *100, 2),
         # derive standard error from margin of error
         perc_employed_se = round(perc_employed_moe / 1.96, 2),
         # derived cv from se
         perc_employed_cv = round((perc_employed_se / perc_employed) * 100, 2)) %>%
  # data was previously grouped by year and age; ungroup data so it is a normal dataframe
  ungroup()
```

```{r}
# line plot of age with Forsyth and all years
age %>%
  filter(geo_description=='Forsyth') %>%
  ggplot(aes(year, perc_employed, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    #facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth County employment rate by age group',
         color = 'Age group',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, perc_employed, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Employment rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
          y = 'Employment Rate',
         x = '') +
    theme_minimal()
```

#### Interpretation

* Employment rate differences by age are stable over time and geographic location.
* Forsyth County's employment rate difference between 25 to 44 and 45 to 64 year old residents is statistically significant in 2016 and 2017.

#### Data

Age data is below.

```{r include = to_include}
age_data <- age %>%
  # select needed variable, reorder to match primary excel data dataframe, and rename to match primary excel dataframe
  select(geo_description, year, age, estimate = perc_employed, moe = perc_employed_moe, 
         se = perc_employed_se, cv = perc_employed_cv)

# add age data to final dataset
tableau_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-age) %>%
    bind_rows(tableau_data, .)
  
age_data %>%
  ff_data_dt(c('Geographic area', 'Year', 'Age group', '% Employed', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('perc_employed', 'perc_employed_moe', c('year', 'geo_description', 'age'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('perc_employed', 'perc_employed_moe', c('year', 'age'))
```

------

### Employment rate by gender

#### Plot

The plots for employment rate by gender are below.

```{r}
# only keep rows related to gender

# gender rows have the following descriptions
gender_descriptions <- c("Employment/Population Ratio; Estimate; Population 20 to 64 years - SEX - Male",
                         "Employment/Population Ratio; Estimate; Population 20 to 64 years - SEX - Female",
                         "Employment/Population Ratio; Estimate; SEX - Male",
                         "Employment/Population Ratio; Estimate; SEX - Female")

gender <- df %>%
  filter(# only keep gender
         description %in% gender_descriptions) %>%
  # create new variable that is only the gender
  # extract either the word Male or Female from the description
  mutate(gender = str_extract(description, 'Male|Female'))

gender %>%
  filter(geo_description == 'Forsyth') %>%
    ggplot(aes(year, estimate, color = gender)) +
      geom_line(size = 1) +
      # create a seperate plot for each geography
      #facet_grid(rows = vars(geo_description)) +
      labs(title = 'Forsyth County employment rate by gender',
           color = 'Gender',
           y = 'Percent employed') +
      scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                         minor_breaks = NULL) +
      theme_minimal()
```

```{r}
gender %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=gender)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Employment rate by gender', ' in ', max(gender$year)),
         fill = 'Gender',
           y = 'Employment Rate',
         x = '') +
    theme_minimal()
```

#### Interpretation

* Males consistently have a higher employment rate than females.
* The Forsyth County difference between the male and female employment is statistically significant for all years except 2013.

#### Data

And here is the underlying data.

```{r include = to_include}
gender_data <- gender %>%
  select(geo_description, year, gender, estimate, moe, se, cv)

# add gender data to final dataset
tableau_data <- gender_data %>%
    # add type of column
    mutate(type = 'Gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Gender', '% Employed', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r include = to_include}
# enter comparison years here
gender_years <- c(max(gender$year), max(gender$year)-1)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'geo_description', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r include = to_include}
gender %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'gender'))
```

-------

### Tableau dataset

This is the final data set that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneded columns
  select(year, geo_description, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
``` 