---
title: "Employment Rate"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs/acs_functions_test.R')

```

The employment rate is the percentage of people age 16 and over who are working. The data comes American Fact Finder table S2301. 

## Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

emp <- read_csv('data/employment_all_years_counties.csv') %>%
  # from counties, remove 'county, North Carolina';
  # this will make it easier to filter based on counties
  mutate(geo_description = str_replace(geo_description, " County, North Carolina", "")) %>%
  # we only need rows representing the employment rate
  # these rows have the phrase 'Employed;' or 'Employment/Population Ratio' in the description
  # change the description so all rows have phrase 'Employment/Population Ratio'; which is the most recent phrase
  mutate(description = str_replace(.$description, 'Employed; ', 'Employment/Population Ratio; ')) %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # only keep employment rate rows and total population counts
         # totals are needed to derive percentages from artificially created groups (age bins)
         str_detect(description, 'Employment[/]Population Ratio;|Total')) %>%
  # the last section of the variable name (label) column is the same for estiamtes and totals
  # this section needs to be extracted so that estimate and total data sets can be merged on it
  mutate(var_name = str_extract(label, 'VC[0-9][0-9]$'))

# make the total counts an additional column

# create new dataset with just totals
emp_total <- emp %>%
  # filter for only totals
  filter(str_detect(description, 'Total'),
         # remove 'PERCENT IMPUTED' rows because these do not reflect counts
         !str_detect(description, 'PERCENT IMPUTED')) %>%
  # only keep the variables that are needed to merge this dataset with the primary dataset
  select(var_name, geo_description, estimate, moe, year)

# remove totals from primary data set and merge in newly created totals data set
emp <- emp %>%
  filter(!str_detect(description, 'Total')) %>%
  left_join(emp_total, by = c('var_name', 'geo_description', 'year')) %>%
  rename(estimate = estimate.x, moe = moe.x,
         population = estimate.y, population_moe = moe.y)
         
```

## US, NC, and county comparisions

### Total employment rate

The plot below shows the overall employment rate of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
emp %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over') %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Employment rate',
         color = 'Comparison Unit') +
    # geom_errorbar(aes(ymin = estimate - (se * 1.96), ymax = estimate + (se * 1.96)), width = 0.2) +
    scale_x_continuous(limits=c(min(emp$year), max(emp$year))) +
    theme_minimal()
```

The table below shows z-scores of all comparison communities for `r max(emp$year)`. To change the comparison year, replace the `max(emp$year)` segment of `year_comparison <- max(emp$year)` with the desired comparison year.

```{r}
year_comparison <- max(emp$year)

emp %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         year == year_comparison) %>%
  ff_acs_zscore('estimate', 'se', 'geo_description') %>%
  kable(caption = paste0("Comparison community z-scores", ' in ', as.character(year_comparison) ))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

### Forsyth County year-to-year trends

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
emp %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore('estimate', 'se', 'year') %>%
  kable(caption = 'Forsyth County yearly trend z-scores') %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

### Forsyth County employment rate by ethnicity

The plot below shows Forsyth County's employment rate by ethnicity ad year. Many hispanic values are missing due to insufficient data.

```{r}
ethnicity_column <- 'description'

keep_ethnicities <- c('White alone, not Hispanic or Latino', 'Black or African American',
                      'Hispanic or Latino origin')

ethnicity <- emp %>%
  # filter for race
  filter(str_detect(description, paste(keep_ethnicities, collapse = '|')),
         geo_description == 'Forsyth') %>%
  # create new column that is only the name of the ethnicity
  mutate(ethnicity = str_extract(description, paste(keep_ethnicities, collapse = '|'))) %>%
  # shorten ethnicity names
  mutate(ethnicity = ifelse(.$ethnicity == 'Black or African American', 'African American',
                            ifelse(.$ethnicity == 'Hispanic or Latino origin', 'Hispanic',
                                   ifelse(.$ethnicity == 'White alone, not Hispanic or Latino', 'White', 'Not sure'))))

ethnicity %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county employment rate by ethnicity',
         color = 'Ethnicity') +
    # geom_errorbar(aes(ymin = estimate - (se * 1.96), ymax = estimate + (se * 1.96)), width = 0.2) +
    scale_x_continuous(limits=c(min(emp$year), max(emp$year))) +
    theme_minimal()
```

Below is a table of z-scores for Forsyth County ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore('estimate', 'se', c('year', 'ethnicity'))  %>%
  kable(caption = 'Forsyth County z-scores by year and ethnicity') %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

## Forsyth County employment rate by age

In examining age, custom age bins were constructed and their employment rates were derived. Bins span the following age groups:

* 16-29
* 30-44
* 45+

The following plot examines Forsyth County's employment rate by age group.

```{r}
# some age bins are artificially created; we will need to derive percentages for these bins

# creat bins for age
age <- emp %>%
  # filter by Forsyth County
  filter(geo_description == 'Forsyth',
         # only keep age
         str_detect(description, ' AGE - '),
         # only keep most recent year
         year == max(year)) %>%
  # create new variable that is only the age
  mutate(age = str_extract(description, 'AGE - .* years'),
         age = str_extract(age, '[0-9].*'),
         # create variable that is the number of people employed in each age bin
         # this is needed to calculate the percentage of people employed in the artificially created age bins
         # must divide estimate by 100 because it is a percentage, but listed as integers
         pop_employed = population * (estimate/100),
         # create function for calculating the moe of two products; will be used below
         # reference: A compass for understaing and using american community survey data, October 2008, A-16
         pop_employed_moe = sqrt((estimate/100)^2 * population_moe^2 + population^2 * (moe/100)^2)) %>%
  # reclassify ages based on new categories
  mutate(age  = ifelse(.$age == "16 to 19 years" | .$age == "20 to 24 years" | .$age == "25 to 29 years", '16-29',
                       ifelse(.$age == "30 to 34 years" | .$age == "35 to 44 years", '30-44', '45+')))

# The following section creates the employment rate and MOE of each artificially derived age bin

age <- age %>%
  # group by year and the newly created age bins
  group_by(year, age) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(population),
            sum_pop_emp = sum(pop_employed),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(population_moe, population),
            sum_pop_emp_moe = moe_sum(pop_employed_moe, pop_employed)) %>% 
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_employed = round( 100 * (sum_pop_emp / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_employed_moe = moe_prop(sum_pop_emp, sum_pop, sum_pop_emp_moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_employed_moe = round( perc_employed_moe *100, 2)) %>% 
  ungroup()

age %>%
  ggplot(aes(age, perc_employed, fill = age)) +
    geom_bar(stat = 'identity') +
    labs(title = 'Forsyth county employment rate by age group',
         y = 'Percent employed') +
    theme_minimal() +
    theme(legend.position="none")

```

The table below highlights z-scores for each age group comparison in `r max(emp$year)`.

```{r}
ff_acs_zscore(age, 'perc_employed', 'perc_employed_moe', 'age')  %>%
  kable(caption = paste0('Forsyth County z-scores by age group', ' in ', as.character(max(age$year)))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```