---
title: "Post secondary Education Completion"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
# load custom chi-square function
source('../../i_education/academic_success/cw_chisq_2way_func.R')
```

3rd grade reading data comes from NC Public Schools website

## Data Import & Transformation/Wrangling

This code block imports and cleans the data. 

```{r}
# create list of raw data files, including path to directory
directory <- 'data/raw_data/'
# create list of raw data files, including path to directory
raw_data_files <- paste0(directory, list.files(directory))

## import each yearly file, and year to dataset, bind with previous years

# initialize dataframe to store results
df <- data.frame()

for (file in raw_data_files) {
  
  # extract the year number from file name
  acad_year <- str_extract(file, '[0-9][0-9][0-9][0-9]-[0-9][0-9]')

  
 # import dataset
  df <- read_delim(file, "\t", escape_double = FALSE, trim_ws = TRUE) %>%
    # add year
    mutate(year = acad_year) %>%
    bind_rows(df, .)
  
}

df_bu <- df
df <- df_bu

# filter out unneeded rows and rename values
df <- df %>%
        filter(school_code=='340LEA'|school_code=='320LEA'|school_code=='410LEA')%>%
        filter(assessment=='ALL')%>%
        filter(subgroup=='FEM'|subgroup=='EDS'|subgroup=='BLCK'|subgroup=='HISP'|subgroup=='MALE'|subgroup=='NOT_EDS'|subgroup=='WHTE'|subgroup=='TOT')%>%
        mutate(name = ifelse(school_code=='320LEA', 'Durham',
                     ifelse(school_code=='410LEA','Guilford','Forsyth')))%>%
        mutate(subgroup = ifelse(subgroup=='FEM','female', ifelse(subgroup=='MALE', 'male',
                                ifelse(subgroup=="BLCK", 'black', ifelse(subgroup=="HISP", 'hispanic', ifelse(subgroup=="WHTE", 'white',
                                        ifelse(subgroup=='EDS','eds',ifelse(subgroup=="NOT_EDS", 'not_eds', 'ALL'))))))))
        
        
        
dfrd <- filter(df, subject=='RD', grade=='03')

dfma <- filter(df, subject=='MA', grade=='08')      

df <- bind_rows(dfrd,dfma)

df <- df %>%
        mutate(subject = ifelse(subject=='RD', '3rd Grade Reading Score', '8th Grade Math Score'))%>%
        select('subject', 'year', 'name', 'subgroup','num_tested','pct_prof')
        

colnames(df) <- c('measure', 'year', 'county', 'group', 'total', 'rate')
       
#Convert variables to numeric type
df$rate <- as.numeric(df$rate)

#Calculate counts for passed and not passed using rate and total
df <- df %>%
        mutate(passed = (rate*total)/100, not_passed = ((100-rate) * total)/100)%>%
        mutate(passed = round(passed,0), not_passed = round(not_passed,0))

rdma2013 <- df
# write out dataset
write_csv(df, 'data/prior_math_reading_scores.csv')

```


```{r}

# combine datasets 
edu <- bind_rows(rdma2018)

# import dataset

#edu <- read_csv('i_education/academic_success/data/prior_math_reading_scores.csv')

#blank data frame for results
comparisons <- data.frame(
  year1=integer(),
  county1=character(),
  group1=character(),
  year2=integer(),
  county2=character(),
  group2=character(),                 
  Yes1=integer(),
  No1=integer(),
  Rate1=double(),
  Yes2=integer(),
  No2=integer(),
  Rate2=double(),
  Significant=logical(),
  Narrative=character(),
  stringsAsFactors=FALSE)

#used to describe results
measure <- "3rd Grade Reading Score"

#list of lists defining what categories to look at
all <- list("ALL")
race_eth <- list('black',"white","hispanic")
gender <- list('male',"female")
EDS <- list('eds',"not_eds")
county1 <- "Forsyth"
county2 <- "Forsyth"
group2 <- "black"
year2 <- "2009-10"

#currently I have data going back to 2007-08

#do the same set of analyses for every demographic group/year combination
demographics <- list(race_eth,gender,EDS,all)
for (year1 in c("2017-18","2016-17","2015-16","2014-15","2013-14","2012-13","2011-12","2010-11","2009-10","2008-09","2007-08")){
  for (demographic_cat in demographics){
    for (group1 in demographic_cat){
      
      
      #there are three different types of analysis
      
        for (year2 in c("2017-18","2016-17","2015-16","2014-15","2013-14","2012-13","2011-12","2010-11","2009-10","2008-09","2007-08")){
                    #Compare this demographic group to itself across years within Forsyth
                    comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year2,"Forsyth",group1))
        }
      
      for (group2 in demographic_cat){
        #Compare this group to every other group in the same demographic cat in this year
        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,"Forsyth",group2))
      }
      
      for (peer in c("Durham","Guilford")){
      #Compare this group to peers in this year 
        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,peer,group1))
      }
    }
  }
}


for (measure in c("8th Grade Math Score")){
        
        #currently I only have FC demographic data going back to 2014-15 and any dat back to 2013-2014
        
        #do the same set of analyses for every demographic group/year combination
        demographics <- list(race_eth,gender,EDS,all)
        for (year1 in c("2017-18","2016-17","2015-16","2014-15","2013-14","2012-13","2011-12","2010-11","2009-10","2008-09","2007-08")){
                for (demographic_cat in demographics){
                        for (group1 in demographic_cat){
                                
                                
                                #there are three different types of analysis
                                
        for (year2 in c("2017-18","2016-17","2015-16","2014-15","2013-14","2012-13","2011-12","2010-11","2009-10","2008-09","2007-08")){
                #Compare this demographic group to itself across years within Forsyth
                                
                #don't make demographic comparisons for 2007-08
                if (year1 != "2007-08" & year2 != "2007-08"){
                        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year2,"Forsyth",group1))
                                        }
                                }
                                
        for (group2 in demographic_cat){
                #Compare this group to every other group in the same demographic cat in this year
                                
                #don't make demographic comparisons for 2013-14
                if (year1 != "2007-08"){
                        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,"Forsyth",group2))
                                        }
                                }
                                
        for (peer in c("Durham","Guilford")){
                                        
                #don't make peer comparisons for 2013-14 except when using overall rates
                if (year1 != "2007-08" | group1 == "ALL"){
                #Compare this group to peers in this year 
                        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,peer,group1))
                                        }
                                }
                                
                        }
                }
        }
}

```

------

### County comparisions

#### Plot

The plot below shows the overall Associate degree rate for the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
reading <- df %>%
        filter(measure == '3rd Grade Reading Score', group =='ALL')

reading %>%
    ggplot(aes(year, rate, color = county)) +
    geom_line(size = 1) +
    labs(title = '3rd Grade Reading Proficiency',
         color = 'County',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```
#### Interpretation

* Not a lot to say here, only 2015-2016 saw a statistically significant year over year change (increase)
* 2017 does not begin to be statistically significant from any other year until 2012.  
* In Forsyth county the drop in 2017 is not statistically significant
* Forsyth and Guilford counties mirror each other/are not SS from each other every year except for 2017

### Associate's degree by ethnicity

#### Plot

The plot below shows Forsyth County's 3rd Grade Reading Scores ethnicity and year. 

```{r fig.width=8,fig.height=10}

ethnicity <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% race_eth))
        
ethnicity %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Ethnicity',
         color = 'Race/Ethnicity',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()


```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Race/Ethnicity',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()

```

#### Interpretation

* White, non-Hispanic rates are statistically significant from African American and Hispanic/Latino rates every year. 
* The African-Americans rate is twice as high as the Hispanic rate.  The White, non-Hispanic rate is 20 percentage points higher than that of African Americans.  And it is more that three times higher than Hispanic/Latino rate. 


### Forsyth County Associate's degree rate by gender

#### Plot

The plot for Associate's degree rate by gender is below.

```{r fig.width=8,fig.height=10}
sex <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% gender))
        
sex %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Gender',
         color = 'Gender',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of gender with all geographies in the most recent year

sex %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Gender',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()


```
#### Interpretation
* Again, though it may look like there are major differences between the genders - no year is statistically significant except for 2014.
* As well, no year is statistically significant from it's predecessor (year over year) for either males or females except for the jump in the female rate between 2015 and 2016, which was sustained in 2017 (as 2017 and 16 were not SS from each other)


```{r}
economic <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% EDS))
        
economic %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Economic Status',
         color = 'Economic Status',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
economic %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency Rate by Economic Status', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Economic Status',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()
```

#### Data

And here is the underlying data.

```{r}
gender_data <- gender %>%
  select(geography, year, gender, estimate, moe, se, cv)

# add gender data to final dataset
tableau_data <- gender_data %>%
    # add type of column
    mutate(type = 'Gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Gender', '% Assoc', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
gender_years <- c(2006:2017)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r}
gender %>%
  filter(geography=='Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'gender'))
```

####Chi-Square test

```{r}

# Run Chi-Square test for statistical significance
# import dataset

edu <- read_csv('i_education/academic_success/data/math_reading_scores.csv')

#blank data frame for results
comparisons <- data.frame(
  year1=integer(),
  county1=character(),
  group1=character(),
  year2=integer(),
  county2=character(),
  group2=character(),                 
  Yes1=integer(),
  No1=integer(),
  Rate1=double(),
  Yes2=integer(),
  No2=integer(),
  Rate2=double(),
  Significant=logical(),
  Narrative=character(),
  stringsAsFactors=FALSE)

#used to describe results
measure <- "3rd Grade Reading Score"

#list of lists defining what categories to look at
all <- list("ALL")
race_eth <- list('black',"white","hispanic")
gender <- list('male',"female")
EDS <- list('eds',"not_eds")

#currently I have data going back to 2013-14

#do the same set of analyses for every demographic group/year combination
demographics <- list(race_eth,gender,EDS,all)
for (year1 in c("2017-18","2016-17","2015-16","2014-15","2013-14")){
  for (demographic_cat in demographics){
    for (group1 in demographic_cat){
      
      
      #there are three different types of analysis
      
        for (year2 in c("2017-18","2016-17","2015-16","2014-15","2013-14")){
                    #Compare this demographic group to itself across years within Forsyth
                    comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year2,"Forsyth",group1))
        }
      
      for (group2 in demographic_cat){
        #Compare this group to every other group in the same demographic cat in this year
        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,"Forsyth",group2))
      }
      
      for (peer in c("Durham","Guilford")){
      #Compare this group to peers in this year 
        comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,peer,group1))
      }
    }
  }
}


for (measure in c("8th Grade Math Score")){
        
        #currently I only have FC demographic data going back to 2014-15 and any dat back to 2013-2014
        
        #do the same set of analyses for every demographic group/year combination
        demographics <- list(race_eth,gender,EDS,all)
        for (year1 in c("2017-18","2016-17","2015-16","2014-15","2013-14")){
                for (demographic_cat in demographics){
                        for (group1 in demographic_cat){
                                
                                
                                #there are three different types of analysis
                                
                                for (year2 in c("2017-18", "2016-17","2015-16","2014-15","2013-14")){
                                        #Compare this demographic group to itself across years within Forsyth
                                        
                                        #don't make demographic comparisons for 2013-14
                                        if (year1 != "2013-14" & year2 != "2013-14"){
                                                comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year2,"Forsyth",group1))
                                        }
                                }
                                
                                for (group2 in demographic_cat){
                                        #Compare this group to every other group in the same demographic cat in this year
                                        
                                        #don't make demographic comparisons for 2013-14
                                        if (year1 != "2013-14"){
                                                comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,"Forsyth",group2))
                                        }
                                }
                                
                                for (peer in c("Durham","Guilford")){
                                        
                                        #don't make peer comparisons for 2013-14 except when using overall rates
                                        if (year1 != "2013-14" | group1 == "ALL"){
                                                #Compare this group to peers in this year 
                                                comparisons <- rbind(comparisons,chisq_2way(measure,year1,"Forsyth",group1,year1,peer,group1))
                                        }
                                }
                                
                        }
                }
        }
}

```
Write out excel workbook if needed.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneeded columns
  select(year, geography, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```