---
title: "3rd Grade Reading Proficiency"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
# load custom chi-square function
source('../../i_education/academic_success/cw_chisq_2way_func.R')
```

3rd grade reading proficiency data comes from NC Public Schools website found here
http://www.ncpublicschools.org/accountability/reporting/leaperformancearchive/

```{r}
df <- read_csv('data/math_reading_scores.csv') %>%
        filter(measure == '3rd Grade Reading Score')%>%
        mutate(year = ifelse(year=='2013-14','2014', 
                             ifelse(year=='2014-15','2015',
                                    ifelse(year=='2015-16','2016',
                                           ifelse(year=='2016-17','2017','2018')))))

df$year <- as.numeric(df$year)

all <- list("ALL")
race_eth <- list('black',"white","hispanic")
gender <- list('male',"female")
EDS <- list('eds',"not_eds")
```

### County comparisions

#### Plot

The plot below shows the 3rd grade reading proficiency rates for the Forsyth, Guilford, and Durham counties.

```{r}

reading <- df

reading %>%
        filter(group =='ALL')%>%
    ggplot(aes(year, rate, color = county)) +
    geom_line(size = 1) +
    labs(title = '3rd Grade Reading Proficiency',
         color = 'County',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(reading$year), max(reading$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```


```{r}
reading %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = county)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency', ' in ', max(reading$year)-1, '-', max(reading$year) ),
       fill = 'County',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()
```

#### Interpretation

* Reading proficiency rates in Forsyth County have decreased over time, 6 percentage points since 2012-14.
* Scores have not changed significantly over the past 3 academic years (16, 17 and 18).  
* Durham's rate is statistically significantly lower from Forsyth County every year.  
* In Forsyth county the drop in 2018 is not statistically significant
* Forsyth and Guilford counties mirror each other/are not SS from each other every year except for 2014-15.

### Reading Proficiency by Ethnicity

#### Plot

The plot below shows Forsyth County's 3rd Grade Reading Proficiency Rates by ethnicity and year. 

```{r fig.width=8,fig.height=10}

ethnicity <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% race_eth))
        
ethnicity %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Ethnicity',
         color = 'Race/Ethnicity',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()


```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Race/Ethnicity',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()

```

#### Interpretation

* African American 3rd grade reading proficiency significantly dropped in 2015-16 and has (statistically) remained at that level since.
* African American students are about half as likely as white students to meet third grade reading proficiency. * Hispanic / Latino students had a similar significant drop between 2013-14 and 2015-16 and has remained at that level since. 
* Over the four years, there are persistent and significant disparities between White, non-Hispanics and both African Americans and Hispanic/Latinos.


### Forsyth County Reading Proficiency rate by gender

#### Plot

The plot for Reading Proficiency rate by gender is below.

```{r fig.width=8,fig.height=10}
sex <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% gender))
        
sex %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Gender',
         color = 'Gender',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of gender with all geographies in the most recent year

sex %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Gender',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()


```
#### Interpretation
* In 2017-18, there was no significant change from 2016-17, males reamined 8% less likely to meet third grade reading proficiency than females.
* Over four years, there is a persistent and significant disparity between between males and females.

### Forsyth County Reading Proficiency rate by economic status

#### Plot

The plot for Reading Proficiency rate by economic status is below.

```{r}
economic <- df %>%
        filter(measure == '3rd Grade Reading Score', (group %in% EDS))
        
economic %>%
        filter(county == 'Forsyth')%>%
        ggplot(aes(year, rate, color = group)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
   labs(title = '3rd Grade Reading Proficiency by Economic Status',
         color = 'Economic Status',
         y = 'Reading Proficiency Rate') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
economic %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(county, rate, fill = group)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('3rd Grade Reading Proficiency Rate by Economic Status', ' in ', max(ethnicity$year)-1, '-', max(ethnicity$year) ),
       fill = 'Economic Status',
       y = 'Reading Proficiency Rate',
       x = '') +
  theme_minimal()
```
#### Interpretation
* The reading proficiencey rate for not economically disadvantaged students continued its downward trend with a significant decrease in 2017-18. 
* Over the four years there are persistent and significant disparities between the two groups of students. 


#### Data

And here is the underlying data.

Write out excel workbook if needed.

```{r}

tableau_data <- df %>%
        select(year, county, measure, group, rate)%>%
        mutate(year = ifelse(year=='2014','2013-14', 
                             ifelse(year=='2015','2014-15',
                                    ifelse(year=='2016','2015-16',
                                           ifelse(year=='2017','2016-17','2017-18')))))%>%
        mutate(group = ifelse(group=='black','African American', 
                      ifelse(group=='white','White, non-Hispanic', 'Hispanic/Latino')))
                                   


colnames(tableau_data) <- c('year', 'geography', 'type', 'subtype', 'estimate')

tableau_data %>%
        # estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
  mutate_at(vars(estimate), funs(./100)) %>%
  # reorder column and do not include unneeded columns
  select(year, geography, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```