---
title: "Demographics: Population by Age"
output:
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs/acs_functions_test.R')

```

The employment rate is the percentage of people age 16 and over who are working. The data comes from U.S. Census Bureau, American Fact Finder table S2301. 

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
# only using Forsyth county for demographics 

df <- read_csv('data/age_all_years.csv')
df <- df %>%
  # convert year to integer for better plotting
    mutate(year = as.integer(year)) %>%
  # from counties, remove 'county, North Carolina';
    mutate(geo_description = str_replace(geo_description, " County, North Carolina", "")) %>%
  # filter for only total population age estimates (remove age by gender rows)
    filter(str_detect(description, "Total;|Percent; Estimate;"))
       
# create 2 dataframes one with only needed age ranges and the other with total population counts
# make the total population an additional column
pop_total <- df %>%
  # filter for only total forsyth county population
  # totals are needed to derive percentages from artificially created groups (age bins)
  filter(str_detect(label,'VC01')) %>%
  # only keep the variables that are needed to merge this dataset with the primary dataset
  select(estimate, description, year)%>%
  filter(!str_detect(description, "Percent;"))

pop_total <- pop_total %>% 
        select(-description)

df_backup <- df
df <- df_backup

df <- df %>%
        filter(str_detect(description, 'Under 5 years|5 to 9|10 to 14|15 to 17|18 to 24|25 to 29|30 to 34|35 to 39|40 to 44|45 to 49|50 to 54|55 to 59|60 to 64|65 to 69|70 to 74|75 to 79|80 to 84|85'))


df <- df [!((df$year == "2017") & (str_detect(df$description, 'Total; Estimate;'))),]


# merge the 2 dataframes
df <- df %>%
  left_join(pop_total, by = 'year') %>%
  rename(estimate = estimate.x,
         total_population = estimate.y)
         
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall age of the Forsyth county.

```{r}
df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over') %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Employment rate',
         color = 'Comparison Unit',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Data

Below is the raw data for employment rates by geographies.

```{r}
df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over') %>%
  select(geo_description, year, estimate, moe, se, cv) %>%
  kable(caption = 'Employment rate by geographic area and year',
        col.names = c('Geographic area', 'Year', '% Employed', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
  
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 
**HOW TO DO ALL YEARS? CAN ESPECIALLY BE USEFUL WHEN COMPARING CURRENT TO PRE-RECESSION, I.E., 2006/07 VS 2017
**OUTPUT XLXS OF ALL YEARS PERHAPS, JUST TRYING TO THINK OF HOW TO MAKE IT EASY TO SEE WHERE "ss" THINGS ARE HAPPENING
**CAN YOU HIGHLIGHT SCORES OVER 1.96?

```{r}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore('estimate', 'se', c('geo_description', 'year')) %>%
  ff_acs_zscore_kable(paste0("Comparison community z-scores", ' in ', paste0(year_comparison, collapse=' and ')))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.
**AGAIN, HIGHLIGHT THOSE #S ABOVE 1.96

```{r}
df %>%
  # only keep total employment rate
  filter(description == 'Employment/Population Ratio; Estimate; Population 16 years and over',
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore('estimate', 'se', 'year') %>%
  ff_acs_zscore_kable('Forsyth County yearly trend z-scores')
```

------

### Forsyth County employment rate by ethnicity

#### Plot

The plot below shows Forsyth County's employment rate by ethnicity and year. Many hispanic values are missing due to insufficient data.

```{r}
ethnicity_column <- 'description'

keep_ethnicities <- c('White alone, not Hispanic or Latino', 'Black or African American',
                      'Hispanic or Latino origin')

ethnicity <- df %>%
  # filter for race
  filter(str_detect(description, paste(keep_ethnicities, collapse = '|')),
         geo_description == 'Forsyth') %>%
  # create new column that is only the name of the ethnicity
  mutate(ethnicity = str_extract(description, paste(keep_ethnicities, collapse = '|'))) %>%
  # shorten ethnicity names
  mutate(ethnicity = ifelse(.$ethnicity == 'Black or African American', 'African American',
                            ifelse(.$ethnicity == 'Hispanic or Latino origin', 'Hispanic',
                                   ifelse(.$ethnicity == 'White alone, not Hispanic or Latino', 'White', 'Not sure'))))

ethnicity %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county employment rate by ethnicity',
         color = 'Ethnicity',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Data

The data for ethnicities is as follows.

```{r}
ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv) %>%
  kable(caption = 'Employment rate by ethnicity',
        col.names = c('Geographic area', 'Year', 'Ethnicity', '% Employed', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

#### Z-scores

Below is a table of z-scores for Forsyth County ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore('estimate', 'se', c('year', 'ethnicity'))  %>%
  ff_acs_zscore_kable('Forsyth County z-scores by year and ethnicity')
```

------

### Forsyth County employment rate by age

#### Plot

The following plot examines Forsyth County's employment rate by age group.

```{r}
# some age bins are artificially created; we will need to derive percentages for these bins

# creat bins for age
age <- df %>%
  # filter by Forsyth County
  filter(geo_description == 'Forsyth',
         # only keep age
         str_detect(description, ' AGE - ')) %>%
         # only keep most recent year
         #year == max(year)) %>%
  # create new variable that is only the age
        mutate(age = str_extract(description, 'AGE - .* years'),
         age = str_extract(age, '[0-9].*'),
         # create variable that is the number of people employed in each age bin
         # this is needed to calculate the percentage of people employed in the artificially created age bins
         # must divide estimate by 100 because it is a percentage, but listed as integers
         pop_employed = population * (estimate/100),
         # create function for calculating the moe of two products; will be used below
         # reference: A compass for understaing and using american community survey data, October 2008, A-16
         pop_employed_moe = sqrt((estimate/100)^2 * population_moe^2 + population^2 * (moe/100)^2)) %>%
  # reclassify ages based on new categories
        
        mutate(age  = ifelse(.$age == "16 to 19 years" | .$age == "20 to 24 years" , '1624',
                       ifelse(.$age == "25 to 29 years" | .$age == "30 to 34 years" | .$age == "35 to 44 years" |
                                .$age == "25 to 44 years", '2544', 
                       ifelse(.$age == "45 to 54 years" | .$age == "55 to 64 years" , '4564','6599'))))

'Under 5 years|5 to 9|10 to 14|15 to 17|18 to 24|25 to 29|30 to 34|35 to 39|40 to 44|45 to 49|50 to 54|55 to 59|60 to 64|65 to 69|70 to 74|75 to 79|80 to 84|85'
# The following section creates the employment rate and MOE of each artificially derived age bin

age <- age %>%
  # group by year and the newly created age bins
  group_by(year, age) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(population),
            sum_pop_emp = sum(pop_employed),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(population_moe, population),
            sum_pop_emp_moe = moe_sum(pop_employed_moe, pop_employed)) %>% 
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_employed = round( 100 * (sum_pop_emp / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_employed_moe = moe_prop(sum_pop_emp, sum_pop, sum_pop_emp_moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_employed_moe = round( perc_employed_moe *100, 2),
         # derive standard error from margin of error
         perc_employed_se = round(perc_employed_moe / 1.96, 2),
         # derived cv from se
         perc_employed_cv = round((perc_employed_se / perc_employed) * 100, 2)) %>% 
  ungroup() #WHAT DOES THIS ACTUALLY DO?

# add Forsyth back as the geographic unit
age$geo_description <- 'Forsyth'

age %>%
    ggplot(aes(year, perc_employed, color = age)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county employment rate by age group',
         color = 'Age group',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

#### Data

Age data is below.

```{r}
age %>%
  select(geo_description, year, age, perc_employed, perc_employed_moe, perc_employed_se, perc_employed_cv) %>%
  kable(caption = 'Employment rate by age',
        col.names = c('Geographic area', 'Year', 'Age group', '% Employed', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore('perc_employed', 'perc_employed_moe', c('year', 'age'))  %>%
  ff_acs_zscore_kable(paste0('Forsyth County z-scores by age group', ' in ', paste(age_years, collapse = ' and ')))
```

------

### Forsyth County employment rate by gender

#### Plot

The plot for employment rate by gender is below.

```{r}
# only keep rows related to Forsyth Count and gender

# gender rows have the following descriptions
gender_descriptions <- c("Employment/Population Ratio; Estimate; Population 20 to 64 years - SEX - Male",
                         "Employment/Population Ratio; Estimate; Population 20 to 64 years - SEX - Female",
                         "Employment/Population Ratio; Estimate; SEX - Male",
                         "Employment/Population Ratio; Estimate; SEX - Female")

gender <- df %>%
  # filter by Forsyth County
  filter(geo_description == 'Forsyth',
         # only keep gender
         description %in% gender_descriptions) %>%
  # create new variable that is only the gender
  # extract either the word Male or Female from the description
  mutate(gender = str_extract(description, 'Male|Female'))

gender %>%
    ggplot(aes(year, estimate, color = gender)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county employment rate by gender',
         color = 'Gender group',
         y = 'Percent employed') +
    scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

#### Data

And here is the underlying data.

```{r}
gender %>%
  select(geo_description, year, gender, estimate, moe, se, cv) %>%
  kable(caption = 'Employment rate by gender',
        col.names = c('Geographic area', 'Year', 'Gender', '% Employed', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
gender_years <- c(max(gender$year), max(gender$year)-1)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore('estimate', 'moe', c('year', 'gender'))  %>%
  ff_acs_zscore_kable(paste0('Forsyth County z-scores by gender', ' in ', paste(gender_years, collapse = ' and ')))
```