---
title: "Median Income Disparities"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(tidycensus)
library(DT)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

Median income disparities data comes from the following ACS tables:

* S1903 (racial disparities by household income)
* B20017 (gender disparities by individual)

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and rows, and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Durham County, NC', 'Forsyth County, NC', 'Guilford County, NC')

df <- read_csv('data/median_income_all_years_counties.csv') %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # remove 'PERCENT IMPUTED' rows
         !str_detect(description, 'PERCENT IMPUTED'),
         # remove 'PERCENT ALLOCATED' rows
         !str_detect(description, 'PERCENT ALLOCATED'),
         # we are only looking at households, so remove families and non-family household
         !str_detect(description, 'FAMILIES -'),
         !str_detect(description, 'NONFAMILY HOUSEHOLDS -'),
         # filter out total columns because we only need median incomes, not counts
         !str_detect(description, 'Total'),
         # starting in 2017, total number of people and percent is in data; remove these so we only have median incomes
         !str_detect(description, '^Number;'),
         !str_detect(description, '^Percent'))

# years will be needed alter
years <- seq(2006, 2017, 1)
```

------

### Median income disparities by race and ethnicity

Median income disparities by ethnicity were derived as follows:

* **White to African American median income disparities:** White, non-Hispanic median income divided by African American median income.
* **White to Hispanic/Latinx median income disparities**: White, non-Hispanic median income divided by Hispanic/Latinx median income.
* **African American to Hispanic/Latinx disparities**: African American median income divided by Hispanic/Latinx median income.
 
#### Create disparities by race and ethnicity dataset

The code block below calculates disparities by ethnicity, along with the se, moe, and cv.

```{r}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description) %>%
  # calculate ratios
  ff_disparities('ethnicity',
                 c('African American', 'Hispanic/Latino', 'White, non-Hispanic'),
                 list(c(3,1), c(3,2), c(1,2)),
                 years, geo_areas)
```

#### Plot

The plots below show median income disparities by ethnicity and year. 

```{r}
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  # add line break to description, which is the legend descrition
  mutate(description = str_replace_all(description, ' to ', ' to\n' )) %>%
  ggplot(aes(year, ratio, color = description)) +
    geom_line(size = 1) +
    # create bold horizontal line at 1
    geom_hline(yintercept = 1) +
    labs(title = 'Forsyth county median income disprities by ethnicity',
         color = 'Comparison',
         y = 'Median income ratio') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  # add line break to description, which is the legend descrition
  mutate(description = str_replace_all(description, ' to ', ' to\n' )) %>%
  ggplot(aes(geo_description, ratio, fill=description)) +
  geom_bar(position = "dodge", stat='identity') +
  # create bold horizontal line at 1
  geom_hline(yintercept = 1) +
  labs(title = paste0('Median income disparity by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Median income ratio',
       x = '') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
```

#### Interpretation

* Within Forsyth County, the White to African American disparity and White to Hispanic/Latinx disparity are similar. In fact, the difference is only statistically significant in 2008.
* Forsyth County's White to African American rate has held steady through the years. The only years with a statistically significant difference from 2017 are 2009 and 2010.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
ethnicity_data <- ethnicity %>%
  select(geo_description, year, description, ratio, moe, se, cv)

# add ethnicity to Tableau dataset
tableau_data <- ethnicity_data %>%
  # add type of column
  mutate(type = 'Race and Ethnicity',
         subtype = description) %>%
  # delete description column since this is now in the subtype column
  select(-description)

ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Comparison', 'Ratio', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below are tables of z scores for disparities within Forsyth County and in all years.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('ratio', 'se', c('year', 'description'))
```

### Median income disparities by gender

Gender income disparities are calculated by dividing female median income by male median income.

```{r include = to_include}
# Median income by gender is in table B20017. This table must be imported and cleaned.
gender <- read_csv('data/median_income_gender_all_years_counties.csv') %>%
  filter(geo_description %in% geo_areas) %>%
  # only include rows for working full time
  filter(str_detect(description,  'Worked full-time')) %>%
  # extract the gender of the row and place in a seperate columns
  mutate(gender = str_extract(description, 'Male|Female'))

gender_F <- gender %>% filter(geo_description == 'Forsyth County, NC')
```

#### Create disparities by gender dataset

The code block below calculates disparities by gender, along with the se, moe, and cv.

```{r}
gender <- gender %>%
  # calculate ratios
  ff_disparities('gender',
                 c('Female', 'Male'),
                 list(c(1, 2)),
                 years, geo_areas)
```

#### Plot

The plots below show median income disparities by gender for all comparison units and years. 

```{r}
gender %>%
  ggplot(aes(year, ratio, color = geo_description)) +
    geom_line(size = 1) +
    # create bold horizontal line at 1
    geom_hline(yintercept = 1) +
    labs(title = 'Median income disprities by gender (female to male)',
         color = 'Comparison',
         y = 'Median income ratio') +
    scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Interpretation

* Forsyth County saw a statistically significant shrinkage in gender income disparity between 2015 and 2016 (disparity rose and became closer to 1). But, the difference between 2015 and 2017 is not statistically significant.

#### Data

The data for gender disparities is as follows.

```{r include = to_include}
gender_data <- gender %>%
  select(geo_description, year, description, ratio, moe, se, cv)

# add ethnicity to Tableau dataset
tableau_data <- gender_data %>%
  # add type of column
  mutate(type = 'Gender',
         subtype = description) %>%
  # delete description column since this is now in the subtype column
  select(-description) %>%
  bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Comparison', 'Ratio', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Here is a tables of z scores for gender disparities within all geographic units in the past two years.

```{r include = to_include}
# enter comparison years here
gender_years <- c(max(gender$year), max(gender$year)-1)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('ratio', 'se', c('geo_description', 'year'))
```

Below are tables of z scores for gender disparities within Forsyth County and in all years.

```{r include = to_include}
gender %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('ratio', 'se', 'year')
```

### Tableau dataset

This is the final data set that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# create table to Tableau data
tableau_data %>%
  select(year, geo_description, type, subtype, ratio) %>%
  # create table
  ff_remove_comparisons() %>%
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```