---
title: "Median Income"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE 
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(tidycensus)
library(DT)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

Median income data comes American Fact Finder table S1903 

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and rows, and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

df <- read_csv('data/median_income_all_years_counties.csv') %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # remove 'PERCENT IMPUTED' rows
         !str_detect(description, 'PERCENT IMPUTED'),
         # remove 'PERCENT ALLOCATED' rows
         !str_detect(description, 'PERCENT ALLOCATED'),
         # we are only looking at households, so remove families and non-family household
         !str_detect(description, 'FAMILIES -'),
         !str_detect(description, 'NONFAMILY HOUSEHOLDS -'),
         # filter out total columns because we only need median incomes, not counts
         !str_detect(description, 'Total'),
         # starting in 2017, total number of people and percent is in data; remove these so we only have median incomes
         !str_detect(description, '^Number;'),
         !str_detect(description, '^Percent'))
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the median income of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
# rows with total median income have the following descriptions
# we will filter on these descriptions to extract rows signifying total median income
total_median_descriptions <- c('Median income (dollars); Estimate; Households',
                               paste0('Median income (dollars); Estimate; HOUSEHOLD INCOME BY RACE AND HISPANIC ', 
                               'OR LATINO ORIGIN OF HOUSEHOLDER - Households'))

df %>%
  # only keep overall median income
  filter(description %in% total_median_descriptions) %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Median income',
         color = 'Comparison Unit',
         y = 'Median income') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```


#### Interpretation

* Forsyth County's estimated median income has generally risen since 2012, but the differences between 2015, 2016, and 2017 are not statistically significant.
* The 2017 difference between Forsyth County's median income and the median income of both North Carolina and Guilford County in 2017 is not statistically significant.

#### Data

Below is the raw data for median income by geographies.

```{r include = to_include}
total_data <- df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions) %>%
  select(geo_description, year, estimate, moe, se, cv)

# this represents the final dataset that will be sent to others for visualizations
tableau_data <- total_data %>%
  # add type of column
  mutate(type = 'Median Income',
         subtype = 'Total Population')

ff_data_dt(total_data,
           c('Geographic area', 'Year', 'Median income', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`.

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Median income by ethnicity

#### Plot

The plot below shows median income by ethnicity and year.

```{r}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description)

ethnicity %>%
  filter(geo_description=='Forsyth') %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    #facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth County median income by ethnicity',
         color = 'Ethnicity') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=ethnicity)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Median income by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Median income',
       x = '') +
  theme_minimal()
```

#### Interpretation

* All ethnicities within Forsyth County have experienced a rise in median income since 2012.
* The difference in median income between African Americans and Hispanic/Latinx is not statistically significant for any year except 2008.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
ethnicity_data <- ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv)

# add ethnicity to Tableau dataset
tableau_data <- ethnicity_data %>%
  # add type of column
  mutate(type = 'Race and Ethnicity',
         subtype = ethnicity) %>%
  # delete ethnicity column since this is now in the subtype column
  select(-ethnicity) %>%
  bind_rows(tableau_data, .)

ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', 'Median Income', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'ethnicity'))
```

A table of z-scores for Forsyth County in all years is below.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'ethnicity'))
```

------

### Median income by age

#### Plot

The following plot examines Median income by age group.

```{r}
age <- df %>%
  # only keep age related rows
  filter(str_detect(description, 'HOUSEHOLD INCOME BY AGE OF HOUSEHOLDER'),
         # remove rows that start with 'Percent Distribution' or 'Number' which is new in 2017
         !str_detect(description, 'Percent Distribution|Number')) %>%
  # extract age from description and place in its own column
  mutate(age = str_extract(description, '[0-9][0-9].*'))
  
  
age %>%
  filter(geo_description == 'Forsyth') %>%
  ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth county median income by age group',
         color = 'Age group') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Median income by age', ' in ', max(age$year)),
         fill = 'Age Group',
         y = 'Median Income',
         x = '') +
    theme_minimal()
```

#### Interpretation

* Within Forsyth County in 2017, the two age groups with the lowest median income saw a decrease in median income, while the two age groups with the highest median income experienced an increase. But, none of these differences are statistically significant.

#### Data

Age data is below.

```{r include = to_include}
age_data <-  age %>%
  select(geo_description, year, age, estimate, moe, se, cv) 

# add age data to tableau dataset
tableau_data <- age_data %>%
  # add type of column
  mutate(type = 'Age Group',
         subtype = age) %>%
  # delete ethnicity column since this is now in the subtype column
  select(-age) %>%
  bind_rows(tableau_data, .)

ff_data_dt(age_data,
           c('Geographic area', 'Year', 'Age group', 'Median income', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'age'))
```

Here are z-scores for all years within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth') %>%
  # shorten age descriptions (take off the word 'year') so they apepar cleaner and shorter as column titles
  mutate(age = str_extract(age, '\\d\\d to \\d\\d|\\d\\d y')) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'age'))
```

------

### Median income by gender

Median income by gender is in table B20017. This table must be imported and cleaned.

```{r}
gender <- read_csv('data/median_income_gender_all_years_counties.csv') %>%
  filter(geo_description %in% geo_areas) %>%
  # only include rows for working full time
  filter(str_detect(description,  'Worked full-time')) %>%
  # extract the gender of the row and place in a seperate columns
  mutate(gender = str_extract(description, 'Male|Female'))
```

#### Plot

The plot for median income by gender is below.

```{r}
gender %>%
  filter(geo_description == 'Forsyth') %>%
    ggplot(aes(year, estimate, color = gender)) +
      geom_line(size=1) +
      # create a seperate plot for each geography
      #facet_grid(rows = vars(geo_description)) +
      labs(title = 'Forsyth County median income by gender',
           color = 'Gender',
           y = 'Median income') +
      scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                         minor_breaks = NULL) +
      theme_minimal()
```

```{r}
gender %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=gender)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Median income by gender', ' in ', max(gender$year)),
         fill = 'Gender',
         y = 'Median Income',
         x = '') +
    theme_minimal()
```

#### Interpretation

* Within Forsyth County, both males and females experienced a general increase in median income since 2006. However, the difference for each gender between any given year and 2017 is not statistically significant.

#### Data

And here is the underlying data.

```{r include = to_include}
gender_data <- gender %>%
  select(geo_description, year, gender, estimate, moe, se, cv)

# add gender data to final dataset
tableau_data <- gender_data %>%
    # add type of column
    mutate(type = 'Gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Gender', '% Employed', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r include = to_include}
# enter comparison years here
gender_years <- c(max(gender$year), max(gender$year)-1)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'geo_description', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r include = to_include}
gender %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'gender'))
```

### Tableau dataset

This is the final data set that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# create table to Tableau data
tableau_data %>%
  select(year, geo_description, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```