---
title: "Median Income"
output:
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(kableExtra)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs/acs_functions_test.R')

```

Median income data comes American Fact Finder table S1903 

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and rows, and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

df <- read_csv('data/median_income_all_years_counties.csv') %>%
  # convert year to integer for better plotting
  mutate(year = as.integer(year)) %>%
  # from counties, remove 'county, North Carolina';
  # this will make it easier to filter based on counties
  mutate(geo_description = str_replace(geo_description, " County, North Carolina", "")) %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # remove 'PERCENT IMPUTED' rows
         !str_detect(description, 'PERCENT IMPUTED'),
         # remove 'PERCENT ALLOCATED' rows
         !str_detect(description, 'PERCENT ALLOCATED'),
         # we are only looking at households, so remove families and non-family household
         !str_detect(description, 'FAMILIES -'),
         !str_detect(description, 'NONFAMILY HOUSEHOLDS -'),
         # filter out total columns because we only need median incomes, not counts
         !str_detect(description, 'Total'))
         
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the median income of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
# rows with total median income have the following descriptions
# we will filter on these descriptions to extract rows signifying total median income
total_median_descriptions <- c('Median income (dollars); Estimate; Households',
                               paste0('Median income (dollars); Estimate; HOUSEHOLD INCOME BY RACE AND HISPANIC ', 
                               'OR LATINO ORIGIN OF HOUSEHOLDER - Households'))

df %>%
  # only keep overall median income
  filter(description %in% total_median_descriptions) %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Median income',
         color = 'Comparison Unit',
         y = 'Median income') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Data

Below is the raw data for median income by geographies.

```{r}
df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions) %>%
  select(geo_description, year, estimate, moe, se, cv) %>%
  kable(caption = 'Median income by geographic area and year',
        col.names = c('Geographic area', 'Year', 'Median income', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
  
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`.

```{r}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore('estimate', 'se', c('geo_description', 'year')) %>%
  kable(caption = paste0("Comparison community z-scores", ' in ', paste0(year_comparison, collapse=' and ') ))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore('estimate', 'se', 'year') %>%
  kable(caption = 'Forsyth County yearly trend z-scores') %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

------

### Forsyth County median income by ethnicity

#### Plot

The plot below shows Forsyth County's median income by ethnicity and year.

```{r}
ethnicity_column <- 'description'

keep_ethnicities <- c('White alone, not Hispanic or Latino', 'Black or African American',
                      'Hispanic or Latino origin')

ethnicity <- df %>%
  # filter for race
  filter(str_detect(description, paste(keep_ethnicities, collapse = '|')),
         # remove rows that start with 'Percent Distribution' or 'Number' which is new in 2017
         !str_detect(description, 'Percent Distribution|Number'),
         geo_description == 'Forsyth') %>%
  # create new column that is only the name of the ethnicity
  mutate(ethnicity = str_extract(description, paste(keep_ethnicities, collapse = '|'))) %>%
  # shorten ethnicity names
  mutate(ethnicity = ifelse(.$ethnicity == 'Black or African American', 'African American',
                            ifelse(.$ethnicity == 'Hispanic or Latino origin', 'Hispanic',
                                   ifelse(.$ethnicity == 'White alone, not Hispanic or Latino', 'White', 'Not sure'))))

ethnicity %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county median income by ethnicity',
         color = 'Ethnicity') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Data

The data for ethnicities is as follows.

```{r}
ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv) %>%
  kable(caption = 'Employment rate by ethnicity',
        col.names = c('Geographic area', 'Year', 'Ethnicity', '% Employed', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

#### Z-scores

Below is a table of z-scores for Forsyth County ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore('estimate', 'se', c('year', 'ethnicity'))  %>%
  kable(caption = 'Forsyth County z-scores by year and ethnicity') %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```

------

### Forsyth County median income by age

#### Plot

The following plot examines Forsyth County's median income by age group.

```{r}
age <- df %>%
  # only keep age related rows
  filter(str_detect(description, 'HOUSEHOLD INCOME BY AGE OF HOUSEHOLDER'),
         geo_description == 'Forsyth',
         # remove rows that start with 'Percent Distribution' or 'Number' which is new in 2017
         !str_detect(description, 'Percent Distribution|Number')) %>%
  # extract age from description and place in its own column
  mutate(age = str_extract(description, '[0-9][0-9].*'))
  
  
age %>%
    ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth county median income by age group',
         color = 'Age group') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Data

Age data is below.

```{r}
age %>%
  select(geo_description, year, age, estimate, moe, se, cv) %>%
  kable(caption = 'Median income by age',
        col.names = c('Geographic area', 'Year', 'Age group', 'Median income', '95% MOE', 'St. Error', 'CV'))  %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years,
         geo_description == 'Forsyth') %>%
  ff_acs_zscore('estimate', 'moe', c('year', 'age'))  %>%
  kable(caption = paste0('Forsyth County z-scores by age group', ' in ', paste(age_years, collapse = ' and '))) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  # bold row names
  column_spec(1, bold = T)
```