---
title: "Median Income"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(tidycensus)
library(DT)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

This markdown file produces individual datasets that filter the data in various ways. If you want to write out the datasets into an excel file where each sheet is a different dataset, change the line of code below to `write_to_excel <- TRUE`.

```{r}
# change to 'TRUE' to write out datasets to excel
write_to_excel <- FALSE
excel_data <- data.frame()
```

Median income data comes American Fact Finder table S1903 

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and rows, and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

df <- read_csv('data/median_income_all_years_counties.csv') %>%
  # only keep rows for the US, NC, Forsyth, Guilford, and Durham
  filter(geo_description %in% geo_areas,
         # remove 'PERCENT IMPUTED' rows
         !str_detect(description, 'PERCENT IMPUTED'),
         # remove 'PERCENT ALLOCATED' rows
         !str_detect(description, 'PERCENT ALLOCATED'),
         # we are only looking at households, so remove families and non-family household
         !str_detect(description, 'FAMILIES -'),
         !str_detect(description, 'NONFAMILY HOUSEHOLDS -'),
         # filter out total columns because we only need median incomes, not counts
         !str_detect(description, 'Total'),
         # starting in 2017, total number of people and percent is in data; remove these so we only have median incomes
         !str_detect(description, '^Number;'),
         !str_detect(description, '^Percent')) %>%
  # change United States to US and North Carolina to NC
  mutate(geo_description = ifelse(.$geo_description == 'United States', 'US',
                                  ifelse(.$geo_description == 'North Carolina', 'NC',
                                         .$geo_description)))
         
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the median income of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
# rows with total median income have the following descriptions
# we will filter on these descriptions to extract rows signifying total median income
total_median_descriptions <- c('Median income (dollars); Estimate; Households',
                               paste0('Median income (dollars); Estimate; HOUSEHOLD INCOME BY RACE AND HISPANIC ', 
                               'OR LATINO ORIGIN OF HOUSEHOLDER - Households'))

df %>%
  # only keep overall median income
  filter(description %in% total_median_descriptions) %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Median income',
         color = 'Comparison Unit',
         y = 'Median income') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_bw()
```

#### Data

Below is the raw data for median income by geographies.

```{r}
total_data <- df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions) %>%
  select(geo_description, year, estimate, moe, se, cv)

if (write_to_excel == TRUE) {
  excel_data <- total_data %>%
    # add type of column
    mutate(type = 'Median Income',
           subtype = 'Total Population')
}

total_data %>%
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
              colnames = c('Geographic area', 'Year', 'Median income', '95% MOE', 'St. Error', 'CV'))) %>%
  # color cv numbers based on value
    formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`.

```{r}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
df %>%
  # only keep total employment rate
  filter(description %in% total_median_descriptions,
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Forsyth County median income by ethnicity

#### Plot

The plot below shows median income by ethnicity and year.

```{r fig.width=8,fig.height=10}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description)

ethnicity %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    facet_grid(rows = vars(geo_description)) +
    labs(title = 'Median income by ethnicity',
         color = 'Ethnicity') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_bw()
```

#### Data

The data for ethnicities is as follows.

```{r}
ethnicity_data <- ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv)

# write out data set to excel only needed
if (write_to_excel == TRUE) {
  
  excel_data <- ethnicity_data %>%
    # add type of column
    mutate(type = 'Race and Ethnicity',
           subtype = ethnicity) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-ethnicity) %>%
    bind_rows(excel_data, .)
}

ethnicity_data %>%
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
            colnames = c('Geographic area', 'Year', 'Ethnicity', 'Median Income', '95% MOE', 'St. Error', 'CV'))) %>%
  # color cv numbers based on value
  formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'ethnicity'))
```

A table of z-scores for Forsyth County in all years is below.

```{r}
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'ethnicity'))
```

------

### Forsyth County median income by age

#### Plot

The following plot examines Median income by age group.

```{r fig.width=8,fig.height=10}
age <- df %>%
  # only keep age related rows
  filter(str_detect(description, 'HOUSEHOLD INCOME BY AGE OF HOUSEHOLDER'),
         # remove rows that start with 'Percent Distribution' or 'Number' which is new in 2017
         !str_detect(description, 'Percent Distribution|Number')) %>%
  # extract age from description and place in its own column
  mutate(age = str_extract(description, '[0-9][0-9].*'))
  
  
age %>%
    ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth county median income by age group',
         color = 'Age group') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_bw()
```

#### Data

Age data is below.

```{r}
age_data <-  age %>%
  select(geo_description, year, age, estimate, moe, se, cv) 

# write out data set to excel only needed
if (write_to_excel == TRUE) {
  
  excel_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-age) %>%
    bind_rows(excel_data, .)
}


age_data %>%
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
              colnames = c('Geographic area', 'Year', 'Age group', 'Median income', '95% MOE', 'St. Error', 'CV'))) %>%
  # color cv numbers based on value
    formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))

```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  # shorten age descriptions (take off the word 'year') so they apepar cleaner and shorter as column titles
  mutate(age = str_extract(age, '\\d\\d to \\d\\d|\\d\\d y')) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'age'))
```

Here are z-scores for all eyars within Forsyth County.

```{r}
age %>%
  filter(geo_description == 'Forsyth') %>%
  # shorten age descriptions (take off the word 'year') so they apepar cleaner and shorter as column titles
  mutate(age = str_extract(age, '\\d\\d to \\d\\d|\\d\\d y')) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'age'))
```

Write out excel workbook if needed.

```{r}
excel_data %>%
  # write to excel
  ff_write_to_excel(., write_to_excel, 'median_income.xlsx')
```