---
title: "Homeownership rate"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

Homeownership data comes from the following U.S. Census Bureau, American Fact Finder tables:

* B25003 (total homeownership)
* B25007 (homeownership by age)
* B25003B (African American homeownership)
* B25003I (Hispanic/Latino homeownership)
* B25003H (White, non-Hispanic homeownership)

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
# create list of geographic areas to keep
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth County, NC', 'Guilford County, NC', 'Durham County, NC')

# import dataset
df <- read_csv('data/homeownership_all_years.csv') %>%
  filter(geo_description %in% geo_areas) %>%
  select(-label)

# age group rows and the other rows have different format so they must be separated
age <- filter(df, file == 'age group')
other <- filter(df, file != 'age group') %>%
  # do not need renter information from this dataset
  filter(!str_detect(description, 'Renter'))

## calculate percentage of households with homes in other

# currently, total households and number of homeowners are on different rows (long form)
# we need both on the same row, in different columns, to calculate percentages

# create separate dataframes for total households and homeowners, 
# then merge these dataframes to create percentiles
other_total <- other %>%
  filter(description == 'Estimate; Total:')
other_home <- other %>%
  filter(description != 'Estimate; Total:')

# merge two datasets, so total households and homeowners are on the same row
# variable name suffixes are '_t' for total households and '_o' for owner
other <- full_join(other_total, other_home, 
                   by = c('geo_description', 'year', 'file' ),
                   suffix = c('_t', '_o'))

# calculate proportions and the se, moe, cv of proportions
# bind to data frame with descriptions of rows
other <- ff_acs_perc(other$estimate_o, other$moe_o, other$estimate_t, other$moe_t) %>%
  bind_cols(other, .) %>%
  # only keep needed columns
  select(geo_description, year, file, prop:cv) %>%
  rename(description = file)
```

```{r}
## calculate percentage of households with homes by age group

# age group homeownership data is formatted differently than the other demographics
# this block calculates the ownership percentages by age

# homeownership percentages are created by dividing the number of homeowners by the number of owners and renters
# we will create function that creates a dataset of just owners or renters
# then, we can combine the datasets and calculate percentages

owner_renter <- function(df, status) {
  
  # create list of age descriptions that will be created
  age_descriptions <- c('15 to 34 years',
                        '35 to 64 years',
                        '65 years and over')

  df %>%
    # only keep values in the needed geographic areas
    filter(geo_description %in% geo_areas,
           # only keep rows for age groups, since other demographics are in previous section
           file == 'age group') %>%
    # remove all rows with descriptions representing renters;
    # only keep rows signifying homeownership
    filter(!str_detect(description, 
                       ifelse(status == 'Owner', 'Renter', 'Owner'))) %>%
    # remove rows that are the total for renters and owners ("Estimate; Total:")
    filter(description != "Estimate; Total:") %>%
    # for age, there is an overall total and a total for all ages and owners
    # remove total for all ages and owners
    filter(!(description == paste0('Estimate; ', status, ' occupied:') & file == 'age group')) %>%
    # change description to only show age group
    mutate(description = str_extract_all(description, ' Householder.*')) %>%
    # delete word household
    mutate(description = str_replace_all(description, 'Householder ', '')) %>%
    # There are too many age group bins in the original data. 
    # They need to be combined into our standard age bins. 
    # The following code block combines age bins.
    mutate(description = str_replace_all(description, '^ 15.*|^ 25.*', age_descriptions[1])) %>%
    mutate(description = str_replace_all(description, '^ 35.*|^ 45.*|^ 55.*|^ 60.*', age_descriptions[2])) %>%
    mutate(description = str_replace_all(description, '^ 65.*|^ 75.*|^ 85.*', age_descriptions[3])) %>%
    # group by the age group and create aggregate counts and MOE for each new age group
    group_by(geo_description, year, description) %>%
    # calculate aggregate counts and SE, ME, and CV for aggregated  age counts
    summarize(estimate = sum(estimate),
              moe = sqrt( sum(moe^2) )) %>%
    mutate(se = moe/1.96,
           cv = (se/estimate) * 100)
}


# create different datasets for owners and renters
owner <- read_csv('data/homeownership_all_years.csv') %>%
  owner_renter('Owner')

renter <- read_csv('data/homeownership_all_years.csv') %>%
  owner_renter('Renter')

# merge owner and renter datasets on year, geography, and description (age)
# variable name suffixes are '_o' for owner and '_r' for renter
df <- full_join(owner, renter, 
                by= c('geo_description', 'year', 'description'), suffix = c("_o", "_r")) %>%
  # create an estimate of the total population for each row by adding the owners and renters
  # this will be used to derive the percentage of homeowners
  mutate(estimate_total = estimate_o + estimate_r,
         # calculate margin of error for the total population in each row
         moe_total = sqrt( moe_o^2 + moe_r^2 ))

# calculate the proportion of homeowners, along with meo, se, and cv
df <- ff_acs_perc(df$estimate_o, df$moe_o, df$estimate_total, df$moe_total) %>%
  bind_cols(df) %>%
  select(geo_description, year, description, prop:cv)

# combine age and other demographic datasets
df <- bind_rows(other, df)

# create type and subtype columns
df <- df %>%
  # Type is age group if there are two numbers at the beginning of the description
  mutate(type = ifelse(str_detect(description, '[0-9][0-9]'), 'Age Group',
                       # type is race/ethnicity if description is one of the racial categories
                       ifelse(str_detect(description, 
                                          'African American|Hispanic[/]Latino|White, non-Hispanic'),
                                          'Race/Ethnicity', 
                              # otherwise, type is the Comparison Community, which is labeled as Comparison Community
                              'Comparison Community'))) %>%
  # change description of 'total' to 'Total Population'
  mutate(description = str_replace_all(description, 'total', 'Total Population')) %>%
  # change column names for conformity with other datasets
  rename(subtype = description, estimate = prop) %>%
  # reorder columns
  select(geo_description, year, type, subtype, estimate, moe, se, cv)
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall homeownership rates for the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
df %>%
  # only keep total employment rate
  filter(type == 'Comparison Community') %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Homeownership rate',
         color = 'Comparison Unit',
         y = 'Homeownership') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Interpretation

* Forsyth County's homeownership rate is higher than Durham County's and Guilford County's rate. The difference between Forsyth and Durham is statistically significant every year. The difference between Forsyth and Guilford is only statistically significant in 2016, 2013, and 2010.
* The homeownership rate in Forsyth County has remained steady in recent years. There is no statistically significant difference in homeownership rates between 2017 and any year until 2009.

#### Data

Below is the raw data for homeownership rates by geographies.

```{r include = to_include}
# this represents the final dataset that will be sent to others for visualizations
tableau_data <- df %>%
  # only keep total employment rate
  filter(type == 'Comparison Community') %>%
  # remove moe, se, and cv
  select(-moe:-cv)

# create table
df %>%
  # only keep total housing rate
  filter(type == 'Comparison Community') %>%
  select(-type, -subtype) %>%
  ff_data_dt(c('Geographic area', 'Year', '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total homeownership rate
  filter(type == 'Comparison Community',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total homeownership rate
  filter(type == 'Comparison Community',
         # only keep data for the most recent year
         geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```


------

### Homeownership rates by ethnicity

#### Plot

The plots below show homeownership rate by ethnicity and year.

```{r}
ethnicity <- df %>%
  # filter for race
  filter(type == 'Race/Ethnicity')

# line plot of ethnicity with Forsyth and all years
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ggplot(aes(year, estimate, color = subtype)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth County homeownership rates by ethnicity',
         color = 'Ethnicity',
         y = 'Homeownership rate (%)') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=subtype)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Homeownership rates by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Homeownership Rate',
       x = '') +
  theme_minimal() +
  theme(axis.text.x=element_text(angle=45,hjust=1))
```

#### Interpretation

* White, non-Hispanics have higher homeownership rates than other ethnicities/races.
* Hispanic/Latinos have almost identical homeownership rates to African Americans. The only year with a statistically significant difference is 2009.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
# add ethnicity information t ofinal datasets
tableau_data <- ethnicity %>%
    select(-moe:-cv) %>%
    bind_rows(tableau_data, .)

ff_data_dt(ethnicity,
           c('Geographic area', 'Year', 'Type', 'Ethnicity', 
             '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'subtype'))
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'subtype'))
```


------

### Homeownership rate by age

#### Plot

The following plots examines the employment rate by age group.

```{r}

age <- filter(df, type == 'Age Group')
# line plot of age with Forsyth and all years
age %>%
  filter(geo_description=='Forsyth County, NC') %>%
  ggplot(aes(year, estimate, color = subtype)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth County homeownership rate by age group',
         color = 'Age group',
         y = 'Homeownership rate (%)') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=subtype)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Homeownership rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
          y = 'Homeownership rate (%)',
         x = '') +
    theme_minimal() +
    theme(axis.text.x=element_text(angle=45,hjust=1))
```

#### Interpretation

* Some age groups have high CV values due to the age group's small sample size. The 15 to 24 year old age group has CV values above 26 every year, and values above 30 most years. The 85 plus age group has CV values above 14 every year.
* Within Forsyth County, the 15 to 24 year old age group has the lowest homeownership rate every year. But, the difference between 15 to 24 year olds and 25 to 34 year olds is not statistically significant in 2016 and 2017.

#### Data

Age data is below.

```{r include = to_include}
# add age data to final dataset
tableau_data <- age %>%
    select(-moe:-cv) %>%
    bind_rows(tableau_data, .)
  
age %>%
  ff_data_dt(c('Geographic area', 'Year', 'Type', 'Age Group', 
               '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'geo_description', 'subtype'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'subtype'))
```

-------

### Tableau dataset

This is the final data set that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  select(year, geo_description, type, subtype, estimate) %>%
  # removes rows containing demographic data for comparison communities 
  # in all years except the latest year
  ff_remove_comparisons() %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
``` 