---
title: "Post secondary Education Completion"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALse}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALse,
                      message = FALse)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')

# also load functions specifically created by Christopher for this indicator, not sure how to do that, but they are # here
#source('../../i_education/post_secondary_completion/post_sec_functions.R')

```

This markdown file produces individual datasets that filter the data in various ways. If you want to write out the datasets into an excel file where each sheet is a different dataset, change the line of code below to `write_to_excel <- TRUE`.

```{r}
# change to 'TRUE' to write out datasets to excel
write_to_excel <- TRUE
```

The post secondary completion rate is the percentage of people age 25 and over with an Associate's Degree or higher. The data comes from U.S. Census Bureau, American Fact Finder table ____. 

### Data Import & dataframe creation 
This block imports the data, calculates the percentages and ses for each estimate and creates a dataframe containing the estimate counts, estimate percentages, and corresponding ses for each geographic location. 


```{r}

edu<-1

#ACS and more intuitive race/ethnic labels
race_eth_label <- c("W","B","H")
race_eth_acs <- c("H","B","I")


for (year in c(2011,2010,2009,2008,2007)){
#for (year in c(2006)){
        
        
        yr2 <- substr(year,3,4)
        
        
        #get totals and gender from B15002
        edu <- acs_newvar(edu,"B15002_001","Pop",age1="25",year=year)
        edu <- acs_newvar(edu,"B15002_002","Pop",age1="25",MF="M",year=year)
        edu <- acs_newvar(edu,"B15002_019","Pop",age1="25",MF="F",year=year)
        edu <- acs_add_sumofvars(edu,c("B15002_014","B15002_015","B15002_016","B15002_017","B15002_018"),"Assoc_p",age1="25",MF="M",year=year)
        edu <- acs_add_sumofvars(edu,c("B15002_031","B15002_032","B15002_033","B15002_034","B15002_035"),"Assoc_p",age1="25",MF="F",year=year)
        
        edu <- acs_add(edu,paste0("Assoc_p2599FA",yr2),paste0("Assoc_p2599MA",yr2),paste0("Assoc_p2599AA",yr2))
        
        edu <- acs_proportion(edu,paste0("Assoc_p2599AA",yr2),paste0("Pop2599AA",yr2),paste0("Pct_Assoc_p2599AA",yr2))
        edu <- acs_proportion(edu,paste0("Assoc_p2599MA",yr2),paste0("Pop2599MA",yr2),paste0("Pct_Assoc_p2599MA",yr2))
        edu <- acs_proportion(edu,paste0("Assoc_p2599FA",yr2),paste0("Pop2599FA",yr2),paste0("Pct_Assoc_p2599FA",yr2))

        
        #race/ethnicity
        
        #in 2008 a separate category for GED was added
        for (i in 1:3){
                edu <- acs_newvar(edu,paste0("B15002",race_eth_acs[i],"_001"),"Pop",age1="25",year=year,race_eth=race_eth_label[i])
                
                if(year>2007){
                        
                        edu <- acs_add_sumofvars(edu,c(paste0("B15002",race_eth_acs[i],"_008"),paste0("B15002",race_eth_acs[i],"_009"),paste0("B15002",race_eth_acs[i],"_010"),paste0("B15002",race_eth_acs[i],"_017"),paste0("B15002",race_eth_acs[i],"_018"),paste0("B15002",race_eth_acs[i],"_019")),"Assoc_p",age1="25",race_eth=race_eth_label[i],year=year)
                }else{
                        edu <- acs_add_sumofvars(edu,c(paste0("B15002",race_eth_acs[i],"_007"),paste0("B15002",race_eth_acs[i],"_008"),paste0("B15002",race_eth_acs[i],"_009"),paste0("B15002",race_eth_acs[i],"_015"),paste0("B15002",race_eth_acs[i],"_016"),paste0("B15002",race_eth_acs[i],"_017")),"Assoc_p",age1="25",race_eth=race_eth_label[i],year=year)
                }
                edu <- acs_proportion(edu,paste0("Assoc_p2599A",race_eth_label[i],yr2),paste0("Pop2599A",race_eth_label[i],yr2),paste0("Pct_Assoc_p2599A",race_eth_label[i],yr2))
        }
        
        
        
        #age
        edu <- acs_add_sumofvars(edu,c("B15001_011","B15001_052"),"Pop",age1="25",age2="34",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_019","B15001_060"),"Pop",age1="35",age2="44",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_027","B15001_068"),"Pop",age1="45",age2="64",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_035","B15001_076"),"Pop",age1="65",age2="99",year=year)
        
        edu <- acs_add_sumofvars(edu,c("B15001_016","B15001_017","B15001_018","B15001_057","B15001_058","B15001_059"),"Assoc_p",age1="25",age2="34",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_024","B15001_025","B15001_026","B15001_065","B15001_066","B15001_067"),"Assoc_p",age1="35",age2="44",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_032","B15001_033","B15001_034","B15001_073","B15001_074","B15001_075"),"Assoc_p",age1="45",age2="64",year=year)
        edu <- acs_add_sumofvars(edu,c("B15001_040","B15001_041","B15001_042","B15001_081","B15001_082","B15001_083"),"Assoc_p",age1="65",age2="99",year=year)
        
        edu <- acs_proportion(edu,paste0("Assoc_p2534AA",yr2),paste0("Pop2534AA",yr2),paste0("Pct_Assoc_p2534AA",yr2))
        edu <- acs_proportion(edu,paste0("Assoc_p3544AA",yr2),paste0("Pop3544AA",yr2),paste0("Pct_Assoc_p3544AA",yr2))
        edu <- acs_proportion(edu,paste0("Assoc_p4564AA",yr2),paste0("Pop4564AA",yr2),paste0("Pct_Assoc_p4564AA",yr2))
        edu <- acs_proportion(edu,paste0("Assoc_p6599AA",yr2),paste0("Pop6599AA",yr2),paste0("Pct_Assoc_p6599AA",yr2))
        
}


names <- colnames(edu)

edu_made <- edu

write.csv(edu_made, "edu_made.csv")
```

## Data Transformation/Wrangling
```{r}
#Change column names to shortened version
colnames(edu_made) <- c('NAME','US',"NC",'Durham','Forsyth','Guilford')

#filter out only rows containing percent data as that is all that is needed
edu_made_pct <- edu_made %>%
        filter(str_detect(NAME, "Pct_Assoc"))

#Gather the geographies to make values for the new varialbe "geography" (in order to conduct z-score test more easily)
tmp <- edu_made_pct %>%
        gather('US',"NC",'Durham','Forsyth','Guilford', key = "geography", value = "estimate")

#Create 2 dataframes out of the original separating out the estimates and the ses 
edu_made_se <- tmp %>%
        filter(str_detect(NAME, "_se"))

edu_made_est <- tmp %>%
        filter(!str_detect(NAME, "_se"))

#Remove "_se" from name to enable merge & change variable name from estimate to se
edu_made_se$NAME <- substr(edu_made_se$NAME,1,19)
edu_made_se <- rename(edu_made_se, se = estimate)

#Merge the two dataframes
edu_attain <- merge(edu_made_est,edu_made_se,by = c("NAME","geography"))

#Round percentages and calculate cv
edu_attain <- edu_attain %>%
        mutate(estimate = 100*(round(estimate,3)), se = 100*(round(se,3)), cv = (round(100*(se/estimate),0)))

#Pull out demographic information from name column, then change the values of those variables to a more descriptive one
edu_attain <- edu_attain %>%
        mutate(age = substr(NAME,12,15), 
               gender = substr(NAME,16,17),
               race_eth = substr(NAME,16,17),
               year = substr(NAME,18,19))

edu_attain$gender <- ifelse(edu_attain$gender == "FA", "female", ifelse(edu_attain$gender == "MA", "male", "ALL"))

edu_attain$race_eth <- ifelse(edu_attain$race_eth == "AW", "White, non-Hispanic", 
                             ifelse(edu_attain$race_eth == "AB", "African American",
                                    ifelse(edu_attain$race_eth == "AH", "Latinx", "ALL")))

edu_attain$year <- paste0("20",edu_attain$year) 

# Change year to numeric class
edu_attain$year <- as.numeric(edu_attain$year)

df <- edu_attain
```


```{r}


peer_comparisons_zscore <- ff_acs_zscore(peer_comparisons,"estimate", "se", var_names = c("geography","year"))

gender <- edu_attain %>%
        filter((geography=='Forsyth'& age=='2599'& race_eth=='ALL' & gender=='Female')|(geography=='Forsyth'& age=='2599'& race_eth=='ALL' & gender=='Male'))

gender_comparisons_zscore <- ff_acs_zscore(gender_comparisons,"estimate", "se", var_names = c("gender","year"))

age <- edu_attain %>%
        filter((geography=='Forsyth'& age=='2534'& race_eth=='ALL' & gender=='ALL')|(geography=='Forsyth'& age=='3544'& race_eth=='ALL' & gender=='ALL')|(geography=='Forsyth'& age=='4564'& race_eth=='ALL' & gender=='ALL')|(geography=='Forsyth'& age=='6599'& race_eth=='ALL' & gender=='ALL'))

age_comparisons_zscore <- ff_acs_zscore(age_comparisons,"estimate", "se", var_names = c("age","year"))

ethnicity <- edu_attain %>%
        filter((geography=='Forsyth' & age=='2599'& race_eth=='White' & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='Hispanic' & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='Black' & gender=='ALL'))

race_eth_comparisons_zscore <- ff_acs_zscore(race_eth_comparisons,"estimate", "se", var_names = c("race_eth","year"))

year_comparisons <- edu_attain %>%
       filter(geography=='Forsyth' & age=='2599'& race_eth=='ALL' & gender=='ALL')

year_comparisons_zscore <- ff_acs_zscore(year_comparisons,"estimate", "se", var_names = "year")


```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall Associate degree rate of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}

df %>%
  # keep all ages, race/ethnicities and genders
  filter(age=='2599',race_eth=='ALL',gender=='ALL') %>%
  ggplot(aes(year, estimate, color = geography)) +
    geom_line(size = 1) +
    labs(title = 'Post secondary Completion',
         color = 'Comparison Unit',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

#### Data

Below is the raw data for Associate degree rate by geographies.

```{r}
peer_comparisons <- peer_comparisons %>%
  select(geography, year, estimate, se, cv)

if (write_to_excel == TRUE) {
  excel_data <- peer_comparisons %>%
    # add type of column
    mutate(type = 'Employment Rate',
           subtype = 'Total Population')
}

# create table
peer_comparisons %>%
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
              colnames = c('Geographic area', 'year', '% Assoc Degree', 'St. Error', 'cv'))) %>%
  # color cv numbers based on value
    formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))





total_data <- df %>%
  # only keep total employment rate
  filter(age=='2599',race_eth=='ALL',gender=='ALL') %>%
  select(geography, year, estimate, se, cv)

# this represents the final dataset that will be sent to others for visualizations
tableau_data <- total_data %>%
    # add type of column
    mutate(type = 'Associate Rate',
           subtype = 'Total Population')

# create table
ff_data_dt(total_data, c('Geography', 'Year', '% Assoc Degree', 'St. Error', 'CV'))
```
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 

```{r}
year_comparison <- c(max(peer_comparisons$year), max(peer_comparisons$year)-1)

peer_comparisons %>%
  # only keep total employment rate
  filter(#description == 'Employment/Population Ratio; estimate; Population 16 years and over',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
 ff_acs_zscore_dt("estimate", "se", var_names = c("geography","year"))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
df %>%
  # only keep Forsyth County, all ages, all race/ethnicities and both genders
  filter(geography=='Forsyth' & age=='2599'& race_eth=='ALL' & gender=='ALL') %>%
         # only keep data for the most recent year
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Associate's degree by ethnicity

#### Plot

The plot below shows Forsyth County's Associate's degree rate by ethnicity and year. Many hispanic values are missing due to insufficient data.

```{r fig.width=8,fig.height=10}

race_eth_comparisons <- edu_attain %>%
        filter((geography=='Forsyth' & age=='2599'& race_eth=='White' & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='Hispanic' & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='Black' & gender=='ALL'))

race_eth_comparisons %>%
  ggplot(aes(year, estimate, color = race_eth)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Percent of Population 25 years and Older with an Associate degree or higher',
         color = 'Ethnicity',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(race_eth_comparisons$year), max(race_eth_comparisons$year), 1),
                       minor_breaks = NULL) +
    theme_bw()


```

#### Data

The data for ethnicities is as follows.

```{r}
ethnicity_data <- ethnicity  %>%
  select(geography, year, race_eth, estimate, se, cv)

# write out data set to excel only needed
if (write_to_excel == TRUE) {
  
  excel_data <- ethnicity_data %>%
    # add type of column
    mutate(type = 'Race and Ethnicity',
           subtype = ethnicity) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-ethnicity) %>%
    bind_rows(excel_data, .)
}

ethnicity_data %>% 
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
            colnames = c('Geographic area', 'year', 'Ethnicity', '% Assoc Degree', 'St. Error', 'cv'))) %>%
  # color cv numbers based on value
  formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'race_eth'))  
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r}
ethnicity %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'race_eth'))
```

------

### Forsyth County Associate degree rate by age

#### Plot

The following plot examines Forsyth County's employment rate by age group.

```{r fig.width=8,fig.height=10}

age %>%
    ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associates Degree or higher rate by age group',
         color = 'age group',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_bw()

```

#### Data

age data is below.

```{r}
age_data <- age %>%
  # select needed variable, reorder to match primary excel data dataframe, and rename to match primary excel dataframe
  select(geography, year, age, estimate, se, cv)

 # write out data set to excel only needed
if (write_to_excel == TRUE) {
  
  excel_data <- age_data %>%
    # add type of column
    mutate(type = 'age Group',
           subtype = age) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-age) %>%
    bind_rows(excel_data, .)
}
  
age_data %>%  
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
              colnames = c('Geographic area', 'year', 'age group', '% Assoc Degree', 'St. Error', 'cv'))) %>%
  # color cv numbers based on value
    formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'age'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r}
age %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('perc_employed', 'perc_employed_moe', c('year', 'age'))
```

------

### Forsyth County Associate's degree rate by gender

#### Plot

The plot for Associate's degree rate by gender is below.

```{r fig.width=8,fig.height=10}
# only keep rows related to gender

gender %>%
    ggplot(aes(year, estimate, color = gender)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associate Degree or Higher rate by gender',
         color = 'gender',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                       minor_breaks = NULL) +
    theme_bw()

```

#### Data

And here is the underlying data.

```{r}
gender_data <- gender %>%
  select(geography, year, gender, estimate, se, cv)

# write out data set to excel only needed
if (write_to_excel == TRUE) {
  
    excel_data <- gender_data %>%
    # add type of column
    mutate(type = 'gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(excel_data, .)
}

gender_data %>%
  datatable(filter='top', extensions='Buttons', 
            options = list(scrollX = TRUE, scrollY = TRUE, dom = 'Bfrtip',
              colnames = c('Geographic area', 'year', 'gender', '% Assoc Degree', 'St. Error', 'cv'))) %>%
  # color cv numbers based on value
    formatStyle('cv', color = styleInterval(c(12, 30), c('black', 'blue', 'red')))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
gender_years <- c(max(gender$year), max(gender$year)-1)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r}
gender %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'gender'))
```


Write out excel workbook if needed.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
excel_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # write to excel
  ff_write_to_excel(., write_to_excel, 'Post_secondary_Completion.xlsx')
```