---
title: "Homeownership rate"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

Homeownership data comes from the following U.S. Census Bureau, American Fact Finder tables:

* B25003 (total homeownership)
* B25007 (homeownership by age)
* B25003B (African American homeownership)
* B25003I (Hispanic/Latinx homeownership)
* B25003H (White, non-Hispanic homeownership)

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
# import dataset
df <- post_sec_all_years

#remove pre-secondary education rows and change label value to a number
df <- df %>% 
        filter(!str_detect(description, 'Less than 9th grade'),(!str_detect(description,'9th to 12th grade')),(!str_detect(description,'High school graduate')),(!str_detect(description,'Some college')), (!str_detect(description,'Regular high school diploma')), (!str_detect(description,'GED or alternative credential'))) %>%
        mutate(label = substr(label,8,9))

df$label <- as.numeric(df$label)


# age group rows and the other rows have different format so they must be separated

#Calculate sum and SEs of total population 25 and over for each geography
#total <- filter(df,file == 'total')
#total <- total %>%
 #       filter(label %in% 11:25 | label %in% 35:49) %>%
  # group by year, geography, and the newly created age bins
  #group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  #summarize(sum_pop = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
  #          sum_pop_moe = moe_sum(moe, estimate))

age <- filter(df, file == 'age group')
white <- filter(df, file == 'White, non-Hispanic')

aa <- filter(df,file == 'African American')
aa_total <- filter(aa, str_detect(description, 'Estimate; Total:'))
aa_male <- filter(aa, str_detect(description, 'Estimate; Male:'))
aa_female <- filter(aa, str_detect(description, 'Estimate; Female:'))
aa <- bind_cols(aa_male,aa_female)

hispanic <- filter(df, file == 'Hispanic/Latino')

male <- filter(df, file == 'gender') %>%
        filter(label %in% 14:18) %>% 
# group by year, geography, and the newly created age bins
  group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
male$gender <- 'male'

female <- filter(df, file == 'gender') %>%
        filter(label %in% 31:35) %>%
        group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
female$gender <- 'female'

gender <- bind_rows(male, female)

gender_total <- filter(df, file == 'gender') %>%
        filter(label==2 | label==19)  
       
gender_total$description <- ifelse(str_detect(gender_total$description, 'Estimate; Female:'), "female", "male")

gender_total <- gender_total %>%
        group_by(year, geo_description, description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(moe, estimate)) %>%
        rename(gender = description)

gender <- left_join(gender,gender_total, by = c('year','geo_description','gender'))

gender <- gender %>%
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (estimate / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(estimate, sum_pop, moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe * 100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2))

gender <- gender %>%
        select(1,2,5,8:11)

#Change column names to shortened version
colnames(gender) <- c('year', 'geography', 'gender', 'estimate', 'moe', 'se', 'cv')



  # data was previously grouped by year and age; ungroup data so it is a normal dataframe
  #ungroup()
pop_total <- filter(df, file == 'gender') %>%
        filter(label==1) %>%  
group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(moe, estimate))
## calculate percentage of households with homes in other

# currently, total households and number of homeowners are on different rows (long form)
# we need both on the same row, in difference columns, to calculate percentages

# create separate dataframes for total households and homeowners, 
# then merge these dataframes to create percentiles
other_total <- other %>%
  filter(description == 'Estimate; Total:')
other_home <- other %>%
  filter(description != 'Estimate; Total:')

# merge two datasets, so total households and homeowners are on the same row
# variable name suffixes are '_t' for total households and '_o' for owner
other <- full_join(other_total, other_home, 
                   by = c('geo_description', 'year', 'file' ),
                   suffix = c('_t', '_o'))

# calculate proportions and the se, moe, cv of proportions
# bind to data frame with descriptions of rows
other <- ff_acs_perc(other$estimate_o, other$moe_o, other$estimate_t, other$moe_t) %>%
  bind_cols(other, .) %>%
  # only keep needed columns
  select(geo_description, year, file, prop:cv) %>%
  rename(description = file)
```

```{r}
## calculate percentage of households with homes by age group

# homeownership percentages are created by dividing the number of homeowners by the number of owners and renters
# we will create function that creates a dataset of just owners or renters
# then, we can combine the datasets and calculate percentages

owner_renter <- function(df, status) {
  df %>%
    # only keep values in the needed geographic areas
    filter(geo_description %in% geo_areas,
           # only keep rows for age groups, since other demographics are in previous section
           file == 'age group') %>%
    # remove all rows with descriptions representing renters;
    # only keep rows signifying homeownership
    filter(!str_detect(description, 
                       ifelse(status == 'Owner', 'Renter', 'Owner'))) %>%
    # remove rows that are the total for renters and owners ("Estimate; Total:")
    filter(description != "Estimate; Total:") %>%
    # for age, there is an overall total and a total for all ages and owners
    # remove total for all ages and owners
    filter(!(description == paste0('Estimate; ', status, ' occupied:') & file == 'age group')) %>%
    # change description to only show age group
    mutate(description = str_extract_all(description, ' Householder.*')) %>%
    # delete word household
    mutate(description = str_replace_all(description, 'Householder ', ''))
}


# create different datasets for owners and renters
owner <- read_csv('data/homeownership_all_years.csv') %>%
  owner_renter('Owner')

renter <- read_csv('data/homeownership_all_years.csv') %>%
  owner_renter('Renter')

# merge owner and renter datasets on year, geography, and description (age)
# variable name suffixes are '_o' for owner and '_r' for renter
df <- full_join(owner, renter, 
                by= c('geo_description', 'year', 'description'), suffix = c("_o", "_r")) %>%
  # create an estimate of the total population for each row by adding the owners and renters
  # this will be used to derive the percentage of homeowners
  mutate(estimate_total = estimate_o + estimate_r,
         # calculate margin of error for the total population in each row
         # use function in tidycensus
         moe_total = sqrt( moe_o^2 + moe_r^2 ))

# calculate the proportion of homeowners, along with meo, se, and cv
df <- ff_acs_perc(df$estimate_o, df$moe_o, df$estimate_total, df$moe_total) %>%
  bind_cols(df) %>%
  select(geo_description, year, description, prop:cv)

# combine age and other demographic datasets
df <- bind_rows(other, df)

# create type and subtype columns
df <- df %>%
  # Type is age group if there are two numbers at the beginning of the description
  mutate(type = ifelse(str_detect(description, '[0-9][0-9]'), 'age group',
                       # type is race/ethnicity if description is one of the racial categories
                       ifelse(str_detect(description, 
                                          'African American|Hispanic[/]Latinx|White, non-Hispanic'),
                                          'race/ethnicity', 
                              # otherwise, type is the total population category
                              'total population'))) %>%
  # change column names for conformity with other datasets
  rename(subtype = description, estimate = prop) %>%
  # reorder columns
  select(geo_description, year, type, subtype, estimate, moe, se, cv)
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall homeownership rates for the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
df %>%
  # only keep total employment rate
  filter(type == 'total population') %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Homeownership rate',
         color = 'Comparison Unit',
         y = 'Homeownership') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

#### Interpretation

* Forsyth County's homeownership rate is higher than Durham County's and Guilford County's rate. The difference between Forsyth and Durham is statistically significant every year. The difference between Forsyth and Guilford is only statistically significant in 2016, 2013, and 2010.
* The homeownership rate in Forsyth County has remained steady in recent years. There is no statistically significant difference in homeownership rates between 2017 and any year until 2009.

#### Data

Below is the raw data for homeownership rates by geographies.

```{r include = to_include}
# this represents the final dataset that will be sent to others for visualizations
tableau_data <- df %>%
  # only keep total employment rate
  filter(type == 'total population') %>%
  # remove moe, se, and cv
  select(-moe:-cv)

# create table
df %>%
  # only keep total housing rate
  filter(type == 'total population') %>%
  select(-type, -subtype) %>%
  ff_data_dt(c('Geographic area', 'Year', '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total homeownership rate
  filter(type == 'total population',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total homeownership rate
  filter(type == 'total population',
         # only keep data for the most recent year
         geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```


------

### Homeownership rates by ethnicity

#### Plot

The plots below show homeownership rate by ethnicity and year.

```{r}
ethnicity <- df %>%
  # filter for race
  filter(type == 'race/ethnicity')

# line plot of ethnicity with Forsyth and all years
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ggplot(aes(year, estimate, color = subtype)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth County homeownership rates by ethnicity',
         color = 'Ethnicity',
         y = 'Homeownership rate (%)') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=subtype)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Homeownership rates by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Homeownership Rate',
       x = '') +
  theme_minimal()
```

#### Interpretation

* White, non-Hispanics have higher homeownership rates than other ethnicities/races.
* Hispanic/Latinos have almost identical homeownership rates to African Americans. The only year with a statistically significant difference is 2009.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
# add ethnicity information t ofinal datasets
tableau_data <- ethnicity %>%
    select(-moe:-cv) %>%
    bind_rows(tableau_data, .)

ff_data_dt(ethnicity,
           c('Geographic area', 'Year', 'Type', 'Ethnicity', 
             '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geo_description', 'subtype'))
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'subtype'))
```


------

### Homeownership rate by age

#### Plot

The following plots examines the employment rate by age group.

```{r}

age <- filter(df, type == 'age group')
# line plot of age with Forsyth and all years
age %>%
  filter(geo_description=='Forsyth') %>%
  ggplot(aes(year, estimate, color = subtype)) +
    geom_line(size = 1) +
    labs(title = 'Forsyth County homeownership rate by age group',
         color = 'Age group',
         y = 'Homeownership rate (%)') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=subtype)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Homeownership rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
          y = 'Homeownership rate (%)',
         x = '') +
    theme_minimal()
```

#### Interpretation

* Some age groups have high CV values due to the age group's small sample size. The 15 to 24 year old age group has CV values above 26 every year, and values above 30 most years. The 85 plus age group has CV values above 14 every year.
* Within Forsyth County, the 15 to 24 year old age group has the lowest homeownership rate every year. But, the difference between 15 to 24 year olds and 25 to 34 year olds is not statistically significant in 2016 and 2017.

#### Data

Age data is below.

```{r include = to_include}
# add age data to final dataset
tableau_data <- age %>%
    select(-moe:-cv) %>%
    bind_rows(tableau_data, .)
  
age %>%
  ff_data_dt(c('Geographic area', 'Year', 'Type', 'Age Group', 
               '% Homeowners', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'geo_description', 'subtype'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'subtype'))
```

-------

### Tableau dataset

This is the final data set that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  select(year, geo_description, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
``` 