---
title: "Post secondary Education Completion"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# prevent printing or R code in the output
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')


```

This markdown file produces individual datasets that filter the data in various ways. If you want to write out the datasets into an excel file where each sheet is a different dataset, change the line of code below to `write_to_excel <- TRUE`.

```{r}
# change to 'TRUE' to write out datasets to excel
write_to_excel <- TRUE
```

The post secondary completion rate is the percentage of people age 25 and over with an Associate's Degree or higher. The data comes from U.S. Census Bureau, American Fact Finder table ____. 

## Data Import & Transformation/Wrangling

```{r}
post_sec <- read_csv('data/raw_data/edu_made.csv')

#Change column names to shortened version
colnames(post_sec) <- c('NAME','US',"NC",'Durham','Forsyth','Guilford')

#filter out only rows containing percent data as that is all that is needed
post_sec_pct <- post_sec %>%
        filter(str_detect(NAME, "Pct_Assoc"))

#Gather the geographies to make values for the new variable "geography" (in order to conduct z-score test more easily)
temp <- post_sec_pct %>%
        gather('US',"NC",'Durham','Forsyth','Guilford', key = "geography", value = "estimate")

#Create 2 dataframes out of the original separating out the estimates and the ses 
post_sec_se <- temp %>%
        filter(str_detect(NAME, "_SE"))

post_sec_est <- temp %>%
        filter(!str_detect(NAME, "_SE"))

#Remove "_se" from name to enable merge & change variable name from estimate to se
post_sec_se$NAME <- substr(post_sec_se$NAME,1,19)
post_sec_se <- rename(post_sec_se, se = estimate)

#Merge the two dataframes
post_sec <- merge(post_sec_est,post_sec_se,by = c("NAME","geography"))

#Change estimate and se to numeric class, round percentages and calculate cv
post_sec$estimate <- as.numeric(post_sec$estimate)
post_sec$se <- as.numeric(post_sec$se)
post_sec <- post_sec %>%
       mutate(estimate = 100*(round(estimate,3)), se = 100*(round(se,3)), cv = (round(100*(se/estimate),0)))

#Pull out demographic information from name column, then change the values of those variables to a more descriptive one
post_sec <- post_sec %>%
        mutate(age = substr(NAME,12,15), 
               gender = substr(NAME,16,17),
               race_eth = substr(NAME,16,17),
               year = substr(NAME,18,19))

post_sec$gender <- ifelse(post_sec$gender == "FA", "female", ifelse(post_sec$gender == "MA", "male", "ALL"))

post_sec$race_eth <- ifelse(post_sec$race_eth == "AW", "White, non-Hispanic", 
                             ifelse(post_sec$race_eth == "AB", "African American",
                                    ifelse(post_sec$race_eth == "AH", "Latinx", "ALL")))

post_sec$year <- paste0("20",post_sec$year) 

# Change year to numeric class
post_sec$year <- as.numeric(post_sec$year)

df <- post_sec
```
------

### NC and county comparisions

#### Plot

The plot below shows the overall Associate degree rate of North Carolina, Forsyth, Guilford, and Durham counties.

```{r}

df %>%
  # keep all ages, race/ethnicities and genders
  filter(age=='2599',race_eth=='ALL',gender=='ALL') %>%
  ggplot(aes(year, estimate, color = geography)) +
    geom_line(size = 1) +
    labs(title = 'Post secondary Completion',
         color = 'Comparison Unit',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```
#### Interpretation

* In Forsyth County, the post secondary education completion rate has steadily increased and every year since 2012 has seen a statistically significant increase year over year. 
* Forsyth County's post secondary completion rate is generally statistically signifcant from all other comparison communities every year with the exception of Guilford County beginning in 2013.

#### Data

Below is the raw data for Associate degree rate by geographies.

```{r}
total_data <- df %>%
  # only keep total employment rate
  filter(age=='2599', race_eth=='ALL', gender=='ALL') %>%
  select(geography, year, estimate, se, cv)

# this represents the final dataset that will be sent to others for visualizations
tableau_data <- total_data %>%
    # add type of column
    mutate(type = 'Associate Rate',
           subtype = 'Total Population over 25')

# create table
ff_data_dt(total_data, c('Geography', 'Year', '% Assoc Degree', 'St. Error', 'CV'))
```
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 


```{r}
year_comparison <- c(2006:2017)

df %>%
  # only keep total employment rate
  filter(age=='2599'& race_eth=='ALL' & gender=='ALL',
         # only keep data for the most recent year
         year %in% year_comparison) %>%
 ff_acs_zscore_dt("estimate", "se", var_names = c("geography","year"))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
df %>%
  # only keep Forsyth County, all ages, all race/ethnicities and both genders
  filter(geography=='Forsyth' & age=='2599'& race_eth=='ALL' & gender=='ALL') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Associate's degree by ethnicity

#### Plot

The plot below shows Forsyth County's Associate's degree rate by ethnicity and year. 

```{r fig.width=8,fig.height=10}

ethnicity <- df %>%
filter((geography=='Forsyth' & age=='2599'& race_eth=="White, non-Hispanic" & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='Latinx' & gender=='ALL')|(geography=='Forsyth' & age=='2599'& race_eth=='African American' & gender=='ALL'))

ethnicity %>%
  ggplot(aes(year, estimate, color = race_eth)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Percent of Population 25 years and Older with an Associate degree or higher',
         color = 'Ethnicity',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()


```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity_bar <- df %>%
filter((age=='2599'& race_eth=="White, non-Hispanic" & gender=='ALL' & year == max(.$year))|(age=='2599'& race_eth=='Latinx' & gender=='ALL' & year == max(.$year))|(age=='2599'& race_eth=='African American' & gender=='ALL' & year == max(.$year)))

ethnicity_bar %>% 
  ggplot(aes(geography, estimate, fill=race_eth)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Assoc degree+ rate by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Assoc Rate',
       x = '') +
  theme_minimal()
```
#### Interpretation

* All race/ethnicities are statistically significant from each other every year. 
* No statistical significance year over year in Hispanic rate Forsyth County's post secondary completion rate is generally statistically signifcant from all other comparison communities every year with the exception of Guilford County beginning in 2013.


#### Data

The data for ethnicities is as follows.

```{r}
ethnicity_data <- ethnicity  %>%
  select(geography, year, race_eth, estimate, se, cv)

# add ethnicity information to final dataset
tableau_data <- ethnicity_data %>%
    # add type of column
    mutate(type = 'Race and Ethnicity',
           subtype = race_eth) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-race_eth) %>%
    bind_rows(tableau_data, .)

ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', '% Assoc', 'St. Error', 'CV'))

```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(2006:2017)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'race_eth'))  
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r}
ethnicity %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'race_eth'))
```

------

### Forsyth County Associate degree rate by age

#### Plot

The following plot examines Forsyth County's Assoc degree rate by age group.

```{r fig.width=8,fig.height=10}

age <- df %>%
        filter((age=='2534'& race_eth=='ALL' & gender=='ALL')|(age=='3544'& race_eth=='ALL' & gender=='ALL')|(age=='4564'& race_eth=='ALL' & gender=='ALL')|(age=='6599'& race_eth=='ALL' & gender=='ALL'))
        
age %>%
    filter(geography=='Forsyth') %>%
    ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associates Degree or higher rate by age group',
         color = 'age group',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
filter(year == max(.$year)) %>%
  ggplot(aes(geography, estimate, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Assoc degree+ rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
          y = 'Assoc degree+ Rate',
         x = '') +
    theme_minimal()
```

#### Data

age data is below.

```{r}
age_data <- age %>%
  # select needed variable, reorder to match primary excel data dataframe, and rename to match primary excel dataframe
  select(geography, year, age, estimate, se, cv)

# add age data to final dataset
tableau_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete age column since this is now in the subtype column
    select(-age) %>%
    bind_rows(tableau_data, .)
  
age_data %>%
  ff_data_dt(c('Geographic area', 'Year', 'Age group', '% Assoc Degree', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r}
# enter comparison years here
age_years <- c(2006:2017)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'age'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r}
age %>%
  filter(geography == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'age'))
```

------

### Forsyth County Associate's degree rate by gender

#### Plot

The plot for Associate's degree rate by gender is below.

```{r fig.width=8,fig.height=10}
# only keep rows related to gender
gender <- df %>%
        filter((age=='2599'& race_eth=='ALL' & gender=='female')|(age=='2599'& race_eth=='ALL' & gender=='male'))

gender %>%
    filter(geography == 'Forsyth') %>%
    ggplot(aes(year, estimate, color = gender)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associate Degree or Higher rate by gender',
         color = 'gender',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of gender with all geographies in the most recent year
gender %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geography, estimate, fill=gender)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Associate Degree or Higher rate by gender', ' in ', max(gender$year)),
         fill = 'Gender',
           y = 'Assoc degree+ Rate',
         x = '') +
    theme_minimal()
```

#### Data

And here is the underlying data.

```{r}
gender_data <- gender %>%
  select(geography, year, gender, estimate, se, cv)

# add gender data to final dataset
tableau_data <- gender_data %>%
    # add type of column
    mutate(type = 'Gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Gender', '% Assoc', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
gender_years <- c(2006:2017)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r}
gender %>%
  filter(geography=='Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'gender'))
```


Write out excel workbook if needed.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneeded columns
  select(year, geography, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```