---
title: "Post secondary Education Completion"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)
```

```{r import}
# import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

Homeownership data comes from the following U.S. Census Bureau, American Fact Finder tables:

* B15001 (Educational attainment by age and sex)
* B15002 (Educational attainment by sex)
* B15002B (African American educational attainment)
* B15002I (Hispanic/Latino educational attainment)
* B15002H (White, non-Hispanic educational attainment)

## Data Import & Transformation/Wrangling

This code block imports and cleans the data. It cleans up variable descriptions.

```{r}
# import dataset & rename counties

df <- read_csv('data/post_sec_all_years.csv')
  df$estimate <- as.numeric(df$estimate)
  df$moe <- as.numeric(df$moe)
  df$se <- as.numeric(df$se)
```

```{r}
age <- filter(df, file == 'age group')
#Males 25 - 34
assoc_m2534 <- filter(age, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, '25 to 34'))

total_m2534 <- filter(assoc_m2534, str_detect(label, 'HD01_VD11'))

assoc_m2534 <- assoc_m2534 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(m2534_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 m2534_assoc_moe = moe_sum(moe, estimate))

#Females 25 - 34
assoc_f2534 <- filter(age, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, '25 to 34'))

total_f2534 <- filter(assoc_f2534, str_detect(label, 'HD01_VD52'))

assoc_f2534 <- assoc_f2534 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(f2534_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 f2534_assoc_moe = moe_sum(moe, estimate))

assoc_2534 <- left_join(assoc_m2534,assoc_f2534, by = c('year','geo_description'))
assoc_2534$age <- "25 to 34 years"

assoc_2534 <- assoc_2534 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_2534_assoc =  m2534_assoc + f2534_assoc,
        sum_2534_assoc_moe = sqrt(f2534_assoc_moe^2 + m2534_assoc_moe^2 ))
      
total_2534 <- left_join(total_m2534,total_f2534, by = c('year','geo_description'), suffix = c("_m", "_f")) 
total_2534 <- total_2534 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_2534_total =  estimate_m + estimate_f,
        sum_2534_total_moe = sqrt(moe_m^2 + moe_f^2 ))

age_2534 <- left_join(assoc_2534,total_2534, by = c('year','geo_description'))

age_2534 <- age_2534 %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_2534_assoc / sum_2534_total), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_2534_assoc, sum_2534_total, sum_2534_assoc_moe, sum_2534_total_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,7,26:29)
age_2534$file <- 'age group'
```


```{r}
age <- filter(df, file == 'age group')
#Males 35 - 44
assoc_m3544 <- filter(age, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, '35 to 44'))

total_m3544 <- filter(assoc_m3544, str_detect(label, 'HD01_VD19'))

assoc_m3544 <- assoc_m3544 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(m3544_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 m3544_assoc_moe = moe_sum(moe, estimate))

#Females 25 - 34
assoc_f3544 <- filter(age, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, '35 to 44'))

total_f3544 <- filter(assoc_f3544, str_detect(label, 'HD01_VD60'))

assoc_f3544 <- assoc_f3544 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(f3544_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 f3544_assoc_moe = moe_sum(moe, estimate))

assoc_3544 <- left_join(assoc_m3544,assoc_f3544, by = c('year','geo_description'))
assoc_3544$age <- "35 to 44 years"

assoc_3544 <- assoc_3544 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_3544_assoc =  m3544_assoc + f3544_assoc,
        sum_3544_assoc_moe = sqrt(f3544_assoc_moe^2 + m3544_assoc_moe^2 ))
      
total_3544 <- left_join(total_m3544,total_f3544, by = c('year','geo_description'), suffix = c("_m", "_f")) 
total_3544 <- total_3544 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_3544_total =  estimate_m + estimate_f,
        sum_3544_total_moe = sqrt(moe_m^2 + moe_f^2 ))

age_3544 <- left_join(assoc_3544,total_3544, by = c('year','geo_description'))

age_3544 <- age_3544 %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_3544_assoc / sum_3544_total), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_3544_assoc, sum_3544_total, sum_3544_assoc_moe, sum_3544_total_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,7,26:29)
age_3544$file <- 'age group'
```

```{r}
age <- filter(df, file == 'age group')
#Males 45 - 64
assoc_m4564 <- filter(age, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, '45 to 64'))

total_m4564 <- filter(assoc_m4564, str_detect(label, 'HD01_VD27'))

assoc_m4564 <- assoc_m4564 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(m4564_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 m4564_assoc_moe = moe_sum(moe, estimate))

#Females 25 - 34
assoc_f4564 <- filter(age, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, '45 to 64'))

total_f4564 <- filter(assoc_f4564, str_detect(label, 'HD01_VD68'))

assoc_f4564 <- assoc_f4564 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(f4564_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 f4564_assoc_moe = moe_sum(moe, estimate))

assoc_4564 <- left_join(assoc_m4564,assoc_f4564, by = c('year','geo_description'))
assoc_4564$age <- "45 to 64 years"

assoc_4564 <- assoc_4564 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_4564_assoc =  m4564_assoc + f4564_assoc,
        sum_4564_assoc_moe = sqrt(f4564_assoc_moe^2 + m4564_assoc_moe^2 ))
      
total_4564 <- left_join(total_m4564,total_f4564, by = c('year','geo_description'), suffix = c("_m", "_f")) 
total_4564 <- total_4564 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_4564_total =  estimate_m + estimate_f,
        sum_4564_total_moe = sqrt(moe_m^2 + moe_f^2 ))

age_4564 <- left_join(assoc_4564,total_4564, by = c('year','geo_description'))

age_4564 <- age_4564 %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_4564_assoc / sum_4564_total), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_4564_assoc, sum_4564_total, sum_4564_assoc_moe, sum_4564_total_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,7,26:29)
age_4564$file <- 'age group'
```


```{r}
age <- filter(df, file == 'age group')
#Males 65
assoc_m6599 <- filter(age, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, '65 years'))

total_m6599 <- filter(assoc_m6599, str_detect(label, 'HD01_VD35'))

assoc_m6599 <- assoc_m6599 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(m6599_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 m6599_assoc_moe = moe_sum(moe, estimate))

#Females 25 - 34
assoc_f6599 <- filter(age, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, '65 years'))

total_f6599 <- filter(assoc_f6599, str_detect(label, 'HD01_VD76'))

assoc_f6599 <- assoc_f6599 %>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(f6599_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 f6599_assoc_moe = moe_sum(moe, estimate))

assoc_6599 <- left_join(assoc_m6599,assoc_f6599, by = c('year','geo_description'))
assoc_6599$age <- "65 years and older"

assoc_6599 <- assoc_6599 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_6599_assoc =  m6599_assoc + f6599_assoc,
        sum_6599_assoc_moe = sqrt(f6599_assoc_moe^2 + m6599_assoc_moe^2 ))
      
total_6599 <- left_join(total_m6599,total_f6599, by = c('year','geo_description'), suffix = c("_m", "_f")) 
total_6599 <- total_6599 %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_6599_total =  estimate_m + estimate_f,
        sum_6599_total_moe = sqrt(moe_m^2 + moe_f^2 ))

age_6599 <- left_join(assoc_6599,total_6599, by = c('year','geo_description'))

age_6599 <- age_6599 %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_6599_assoc / sum_6599_total), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_6599_assoc, sum_6599_total, sum_6599_assoc_moe, sum_6599_total_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,7,26:29)
age_6599$file <- 'age group'
```


```{r}

aa <- filter(df,file == 'African American')
aa_total <- filter(aa, str_detect(description, 'Estimate; Total:'))
aa_male <- filter(aa, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(bm_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 bm_assoc_moe = moe_sum(moe, estimate))
aa_female <- filter(aa, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(bf_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 bf_assoc_moe = moe_sum(moe, estimate))
        
aa_all <- left_join(aa_male,aa_female, by = c('year','geo_description'))

aa_all <- aa_all %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_b_assoc = bf_assoc + bm_assoc,
        sum_b_assoc_moe = sqrt(bf_assoc_moe^2 + bm_assoc_moe^2 ))
      
aa_all <- left_join(aa_all,aa_total, by = c('year','geo_description'))   

aa_prop <- aa_all %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_b_assoc / estimate), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_b_assoc, estimate, sum_b_assoc_moe, moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,15:19)
  # data was previously grouped by year and age; ungroup data so it is a normal dataframe

```

```{r}
hispanic <- filter(df, file == 'Hispanic/Latino')
hispanic_total <- filter(hispanic, str_detect(description, 'Estimate; Total:'))
hispanic_male <- filter(hispanic, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(hm_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 hm_assoc_moe = moe_sum(moe, estimate))
hispanic_female <- filter(hispanic, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(hf_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 hf_assoc_moe = moe_sum(moe, estimate))
        
hispanic_all <- left_join(hispanic_male,hispanic_female, by = c('year','geo_description'))

hispanic_all <- hispanic_all %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_h_assoc = hf_assoc + hm_assoc,
        sum_h_assoc_moe = sqrt(hf_assoc_moe^2 + hm_assoc_moe^2 ))
      
hispanic_all <- left_join(hispanic_all,hispanic_total, by = c('year','geo_description'))   

hispanic_prop <- hispanic_all %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_h_assoc / estimate), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_h_assoc, estimate, sum_h_assoc_moe, moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,15:19)
```

```{r}
white <- filter(df, file == 'White, non-Hispanic')
white_total <- filter(white, str_detect(description, 'Estimate; Total:'))
white_male <- filter(white, str_detect(description, 'Estimate; Male:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(wm_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 wm_assoc_moe = moe_sum(moe, estimate))
white_female <- filter(white, str_detect(description, 'Estimate; Female:'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Graduate'))%>%
        group_by(year, geo_description) %>%
                summarize(wf_assoc = sum(estimate),
                 # deriving MOE from aggregagate count data using tidycensus package
                 wf_assoc_moe = moe_sum(moe, estimate))
        
white_all <- left_join(white_male,white_female, by = c('year','geo_description'))

white_all <- white_all %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_w_assoc = wf_assoc + wm_assoc,
        sum_w_assoc_moe = sqrt(wf_assoc_moe^2 + wm_assoc_moe^2 ))
      
white_all <- left_join(white_all,white_total, by = c('year','geo_description'))   

white_prop <- white_all %>%         
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_w_assoc / estimate), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_w_assoc, estimate, sum_w_assoc_moe, moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe *100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2)) %>%
        select(1,2,15:19)
```



```{r}

female <- filter(df, file == 'gender') %>%
        filter(str_detect(description, 'Female'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Master|Professional|Doctorate'))%>%
        group_by(year, geo_description) %>%
        # find the total population and total employed population in artificially derived age bins
        # must find total populations and total employed so percentages can be reconstructed
        summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
female$gender <- 'female'

male <- filter(df, file == 'gender') %>%
        filter(!str_detect(description, 'Female'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Master|Professional|Doctorate'))%>%
  group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
male$gender <- 'male'

gender <- bind_rows(male, female)

gender_total <- filter(df, file == 'gender') %>%
        filter(label=='HD01_VD19' | label=='HD01_VD02')  
       
gender_total$description <- ifelse(str_detect(gender_total$description, 'Estimate; Female:'), "female", "male")

gender_total <- gender_total %>%
        group_by(year, geo_description, description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(moe, estimate)) %>%
        rename(gender = description)

gender <- left_join(gender,gender_total, by = c('year','geo_description','gender'))

gender <- gender %>%
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (estimate / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(estimate, sum_pop, moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe * 100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2))

gender <- gender %>%
        select(1,2,5,8:11)

#Change column names to shortened version
colnames(gender) <- c('year', 'geography', 'gender', 'estimate', 'moe', 'se', 'cv')

```

```{r}
f_assoc <- filter(df, file == 'gender') %>%
        filter(str_detect(description, 'Female'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Master|Professional|Doctorate'))%>%
        group_by(year, geo_description) %>%
        # find the total population and total employed population in artificially derived age bins
        # must find total populations and total employed so percentages can be reconstructed
        summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
f_assoc$gender <- 'female'

m_assoc <- filter(df, file == 'gender') %>%
        filter(!str_detect(description, 'Female'))%>%
        filter(str_detect(description, 'Associate|Bachelor|Master|Professional|Doctorate'))%>%
  group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(estimate = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            moe = moe_sum(moe, estimate))
m_assoc$gender <- 'male'

assoc_mf <- left_join(m_assoc, f_assoc, by = c('year','geo_description'), suffix = c('_m','_f'))

assoc_mf <- assoc_mf %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  mutate(sum_assoc = estimate_m + estimate_f,
        sum_assoc_moe = sqrt(moe_m^2 + moe_f^2 ))
      
pop25_total <- filter(df, file == 'gender') %>%
        filter(label=='HD01_VD19' | label=='HD01_VD02')  
       
pop25_total <- pop25_total %>%
        group_by(year, geo_description) %>%
  # find the total population and total employed population in artificially derived age bins
  # must find total populations and total employed so percentages can be reconstructed
  summarize(sum_pop = sum(estimate),
            # deriving MOE from aggregagate count data using tidycensus package
            sum_pop_moe = moe_sum(moe, estimate))

assoc_all <- left_join(assoc_mf,pop25_total, by = c('year','geo_description'))   

assoc_all <- assoc_all %>%
  # calculate the derived proportion of the employment rate for each age bin
  # multiple estimate by 100 so it is in same percentile format as original data
  # round to two decimal places
  mutate(perc_assoc = round( 100 * (sum_assoc / sum_pop), 2),
         # calculate MOE from derived proportion using tidycensus
         perc_assoc_moe = moe_prop(sum_assoc, sum_pop, sum_assoc_moe, sum_pop_moe),
         # multiply by 100 and round to 2 decimals for consistend formatting
         perc_assoc_moe = round(perc_assoc_moe * 100, 2),
         # derive standard error from margin of error
         perc_assoc_se = round(perc_assoc_moe / 1.96, 2),
         # derived cv from se
         perc_assoc_cv = round((perc_assoc_se / perc_assoc) * 100, 2))

assoc_all <- assoc_all %>%
        select(1,2,13:16)

#Change column names to shortened version
colnames(assoc_all) <- c('year', 'geography', 'estimate', 'moe', 'se', 'cv')

```

------

### US, NC and county comparisions

#### Plot

The plot below shows the overall Associate degree rate for the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}

assoc_all %>%
    ggplot(aes(year, estimate, color = geography)) +
    geom_line(size = 1) +
    labs(title = 'Post secondary Completion',
         color = 'Comparison Unit',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```
#### Interpretation

* Not a lot to say here, only 2015-2016 saw a statistically significant year over year change (increase)
* 2017 does not begin to be statistically significant from any other year until 2012.  
* The drop in 2017 is not statistically significant

#### Data

Below is the raw data for Associate degree rate by geographies.

```{r}
total_data <- assoc_all %>%
  select(geography, year, estimate, moe, se, cv)

# this represents the final dataset that will be sent to others for visualizations
tableau_data <- total_data %>%
    # add type of column
    mutate(type = 'Associate Rate',
           subtype = 'Comparison Community')

# create table
ff_data_dt(total_data, c('Geography', 'Year', '% Assoc Degree', '95% MOE', 'St. Error', 'CV'))

# duplicate dataset of just Forsyth County with subtype "Total Population"
forsyth_data <- assoc_all %>%
  filter(geography=="Forsyth County, NC")%>%
  select(geography, year, estimate, moe, se, cv)

tableau_data <- forsyth_data %>%
    # add type of column
    mutate(type = 'Associate Rate',
           subtype = 'Total Population')
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`. 


```{r}
year_comparison <- c(2006:2017)

assoc_all %>%
  # only keep total employment rate
  filter(year %in% year_comparison) %>%
 ff_acs_zscore_dt("estimate", "se", var_names = c("geography","year"))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r}
assoc_all %>%
  # only keep Forsyth County
  filter(geography=='Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Associate's degree by ethnicity

#### Plot

The plot below shows Forsyth County's Associate's degree rate by ethnicity and year. 

```{r fig.width=8,fig.height=10}

ethnicity <- bind_rows(aa_prop,white_prop,hispanic_prop) 
colnames(ethnicity) <- c('year','geography','race','estimate','moe','se','cv')

ethnicity %>%
        filter(geography=="Forsyth County, NC")%>%
        ggplot(aes(year, estimate, color = race)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Percent of Population 25 years and Older with an Associate degree or higher',
         color = 'Ethnicity',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()


```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>% 
  filter(year == max(.$year))%>%
  ggplot(aes(geography, estimate, fill = race)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Assoc degree+ rate by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Race/Ethnicity',
       y = 'Assoc Rate',
       x = '') +
  theme_minimal()
```

#### Interpretation

* African American and White, non-Hispanic rates are statistically significant from each other every year. 
* I need to look into the Hispanic rates.  The program calculated rates based on counts that were given, but those counts did not come with MOEs so the data is not reliable.  Need to discuss with Shane.


#### Data

The data for ethnicities is as follows.

```{r}
ethnicity_data <- ethnicity  %>%
  select(geography, year, race, estimate, moe, se, cv)

# add ethnicity information to final dataset
tableau_data <- ethnicity_data %>%
    # add type of column
    mutate(type = 'Race and Ethnicity',
           subtype = race) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-race) %>%
    bind_rows(tableau_data, .)

ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', '% Assoc', '95% MOE', 'St. Error', 'CV'))

```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r}
# enter comparison years here
ethnicity_years <- c(2006:2017)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'race'))  
```

Below is a table of z-scores for ethnicities in all years within Forsyth County.

```{r}
ethnicity %>%
  filter(geography=="Forsyth County, NC")%>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'race'))
```

------

### Forsyth County Associate degree rate by age

#### Plot

The following plot examines Forsyth County's Assoc degree rate by age group.

```{r fig.width=8,fig.height=10}

age <- bind_rows(age_2534,age_3544,age_4564,age_6599)
colnames(age) <- c('year','geography','age','estimate','moe','se','cv','file')

age %>%
    filter(geography=='Forsyth County, NC') %>%
    ggplot(aes(year, estimate, color = age)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associates Degree or higher rate by age group',
         color = 'Age Group',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()

```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
filter(year == max(.$year)) %>%
  ggplot(aes(geography, estimate, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Assoc degree+ rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
          y = 'Assoc degree+ Rate',
         x = '') +
    theme_minimal()
```
#### Interpretation

* Even thought it looks like soemthing really dramatic is happening with 654 years and older folks, there has been no statistically significant increase year over year.  Though it has certainly been rising over time.  A jump of 
12 percentage points from 2008 to 2017.

#### Data

age data is below.

```{r}
age_data <- age %>%
  # select needed variable, reorder to match primary excel data dataframe, and rename to match primary excel dataframe
  select(geography, year, age, estimate, moe, se, cv)

# add age data to final dataset
tableau_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete age column since this is now in the subtype column
    select(-age) %>%
    bind_rows(tableau_data, .)
  
age_data %>%
  ff_data_dt(c('Geographic area', 'Year', 'Age group', '% Assoc Degree', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group and geography in the past two years.

```{r}
# enter comparison years here
age_years <- c(2006:2017)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'age'))
```

Below is a table of z-scores for all years and ages within Forsyth County.

```{r}
age %>%
  filter(geography == 'Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'age'))
```

------

### Forsyth County Associate's degree rate by gender

#### Plot

The plot for Associate's degree rate by gender is below.

```{r fig.width=8,fig.height=10}

gender %>%
    filter(geography == 'Forsyth County, NC') %>%
    ggplot(aes(year, estimate, color = gender)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    labs(title = 'Associate Degree or Higher rate by gender',
         color = 'Gender',
         y = '% Pop 25 years & older with Associate Degree or Higher') +
    scale_x_continuous(breaks = seq(min(gender$year), max(gender$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of gender with all geographies in the most recent year
gender %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geography, estimate, fill=gender)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Associate Degree or Higher rate by gender', ' in ', max(gender$year)),
         fill = 'Gender',
           y = 'Assoc degree+ Rate',
         x = '') +
    theme_minimal()
```
#### Interpretation
* Again, though it may look like there are major differences between the genders - no year is statistically significant except for 2014.

#### Data

And here is the underlying data.

```{r}
gender_data <- gender %>%
  select(geography, year, gender, estimate, moe, se, cv)

# add gender data to final dataset
tableau_data <- gender_data %>%
    # add type of column
    mutate(type = 'Gender',
           subtype = gender) %>%
    # delete gender column since this is now in the subtype column
    select(-gender) %>%
    bind_rows(tableau_data, .)

ff_data_dt(gender_data,
           c('Geographic area', 'Year', 'Gender', '% Assoc', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r}
# enter comparison years here
gender_years <- c(2006:2017)

gender %>%
  filter(year %in% gender_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'geography', 'gender'))
```

And here are Forsyth County's z-scores for all years.

```{r}
gender %>%
  filter(geography=='Forsyth County, NC') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'gender'))
```


Write out excel workbook if needed.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneeded columns
  select(year, geography, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```