---
title: "Poverty Rates"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
# set value of whether the file is for an analyst, and therefore all resutls and code blocks are needed,
# or whether the file is for another, and some results and all code blocks are hidden
for_analyst <- TRUE

# global options for code blocks
knitr::opts_chunk$set(warning = FALSE, # supress warnings
                      message = FALSE, # supress messages
                      # do not print out code blocks if for_analyst equals FALSE 
                      echo = FALSE)
                      #echo = ifelse(for_analyst == FALSE, FALSE, TRUE))

# value that sets the 'include' parameter in Markdown based on whether the document is for the analyst
to_include = ifelse(for_analyst == TRUE, TRUE, FALSE)                     
```

```{r import}
 # import functions and packages
library(tidyverse)
library(knitr)
library(DT)
library(tidycensus)
# load custom ACS functions
source('../../functions/acs_load_funcs.R')
```

The poverty rate reflects the percentage of the population in poverty. The data for poverty rates comes American Fact Finder table S1701. For poverty counts of smaller age groups go to B17001.

### Data cleaning

This code block imports and cleans the data. It filters out unneeded NC counties and cleans up variable descriptions.

```{r}
geo_areas <- c('United States', 'North Carolina', 
                'Forsyth', 'Guilford', 'Durham')

df <- read_csv('data/poverty_all_years_counties.csv') %>%
  # only keep selected NC counties, the state of NC, and US data
  filter(geo_description %in% geo_areas,
         # only need rows showing the percent below poverty level, don't need count data
         str_detect(description, '^Percent'))
```

------

### US, NC, and county comparisions

#### Plot

The plot below shows the overall employment rate of the US, North Carolina, and Forsyth, Guilford, and Durham counties.

```{r}
df %>%
  # only keep total poverty rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$')) %>%
  ggplot(aes(year, estimate, color = geo_description)) +
    geom_line(size = 1) +
    labs(title = 'Poverty rate by geographic area',
         color = 'Comparison Unit',
         y = 'Poverty rate (%)') +
    scale_x_continuous(breaks = seq(min(df$year), max(df$year), 1)) +
    theme_minimal()
```

#### Interpretation

* In 2017, Forsyth County had the highest poverty rate among comparison geographic units. But, the difference between Forsyth and Durham counties is not statistically significant.
* Forsyth County's poverty rates have consistently dropped since 2012. The differences in poverty rates between 2017 and 2012, 2013, and 2014 are statistically significant. 

#### Data

Below is the raw data for poverty rates by geographies.

```{r include = to_include}
total_data <- df %>%
  # only keep total poverty rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$')) %>%
  select(geo_description, year, estimate, moe, se, cv)

# create dataset formatted for Tableau
tableau_data <- total_data %>%
  # add type of column
  mutate(type = 'Poverty Rate',
         subtype = 'Total Population')

# create table
ff_data_dt(total_data, c('Geographic area', 'Year', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below shows z-scores of all comparison communities for the past two years. To change the comparison year, replace the `c(max(df$year), max(df$year)-1)` segment of `year_comparison <- max(df$year)` with the desired year or vector of years such as `c(2013, 2014, 2015)`.

```{r include = to_include}
year_comparison <- c(max(df$year), max(df$year)-1)

df %>%
  # only keep total employment rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$'),
         # only keep data for the most recent year
         year %in% year_comparison) %>%
  ff_acs_zscore_dt('estimate', 'se', c('geo_description', 'year'))
```

------

### Forsyth County year-to-year trends

#### Z-scores

The plot above shows Forsyth County's yearly trend and the table below lists yearly z-scores.

```{r include = to_include}
df %>%
  # only keep total employment rate
  filter(str_detect(description, '^Percent *below poverty level; Estimate; Population for whom poverty status is determined$'),
         # only keep Forsyth
         geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', 'year')
```

------

### Poverty rate by ethnicity

#### Plot

The plot below shows poverty rates by ethnicity and year.

```{r}
ethnicity <- df %>%
  # filter for race
  ff_acs_ethnicity(description)

ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ggplot(aes(year, estimate, color = ethnicity)) +
    geom_line(size = 1) +
    # create a seperate plot for each geography
    #facet_grid(rows = vars(geo_description)) +
    labs(title = 'Forsyth County poverty rates by ethnicity',
         color = 'Ethnicity',
         y = 'Poverty rate (%)') +
    scale_x_continuous(breaks = seq(min(ethnicity$year), max(ethnicity$year), 1),
                       minor_breaks = NULL) +
    theme_minimal()
```

```{r}
# bar chart of ethnicity with all geographies in the most recent year
ethnicity %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=ethnicity)) +
  geom_bar(position = "dodge", stat='identity') +
  labs(title = paste0('Poverty rate by ethnicity', ' in ', max(ethnicity$year)),
       fill = 'Ethnicity',
       y = 'Poverty rate (%)',
       x = '') +
  theme_minimal()
```

#### Interpretation

* The African American poverty rate has fallen 10 percentage points since 2012.
* THe White, non-Hispanic poverty rate has remained stable over the pst 10 years. Over this time period, no years have statistically significant differences from 2017.
* Most of the Hispanic/Latinx values have high CV scores (over 12).  Therefore, our confidence in Hispanic/Latinx estimates is weak.

#### Data

The data for ethnicities is as follows.

```{r include = to_include}
ethnicity_data <- ethnicity %>%
  select(geo_description, year, ethnicity, estimate, moe, se, cv)

# add to dataset formatted for Tableau
tableau_data <- ethnicity_data %>%
  # add type of column
  mutate(type = 'Race and Ethnicity',
         subtype = ethnicity) %>%
  # delete ethnicity column since this is now in the subtype column
  select(-ethnicity) %>%
  bind_rows(tableau_data, .)
  
ff_data_dt(ethnicity_data,
           c('Geographic area', 'Year', 'Ethnicity', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

Below is a table of z-scores for ethnicities in the past two years. To generate z-scores for other years, replace `c(max(ethnicity$year), max(ethnicity$year)+1)` in `ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)+1)` with the desired year.

```{r include = to_include}
# enter comparison years here
ethnicity_years <- c(max(ethnicity$year), max(ethnicity$year)-1)

ethnicity %>%
  filter(year %in% ethnicity_years) %>%
  ff_acs_zscore_dt('estimate', 'se', c('year','geo_description', 'ethnicity'))
```

The next table has ethnicity z-scores for all years within Forsyth County.

```{r include = to_include}
ethnicity %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'se', c('year', 'ethnicity'))

```

------

### Poverty by age

#### Plot

In 2015, the Census created additional age groups, such as under 5 and 5 to 17. Prior to 2015, however, the youngest age group was 18 and under. Therefore, yearly age comparisons only trace back to 2015.

```{r}
# list of the end of age rows, to filter on
age_descriptions <- c('Under 5 years',
                      '5 to 17 years',
                      '18 to 34 years',
                      '35 to 64 years',
                      '65 years and over')

# create regular expression to filter data set and only keep rows that have one of the 
# age descriptions at the end of the row description
age_re <- paste0(age_descriptions, collapse = '$|')

age <- df %>%
  # only keep variables for 2015 and later
  filter(year >= 2015,
         str_detect(description, age_re)) %>%
  # create new column that extracts the age from the description column
  mutate(age = str_extract(description, age_re))
  
age %>%
    filter(geo_description =='Forsyth') %>%
    ggplot(aes(year, estimate, color = age)) +
      geom_line(size = 1) +
      # create a seperate plot for each geography
      #facet_grid(rows = vars(geo_description)) +
      labs(title = 'Forsyth County poverty rate by age group',
           color = 'Age group',
           y = 'Poverty rate (%)') +
      scale_x_continuous(breaks = seq(min(age$year), max(age$year), 1),
                         minor_breaks = NULL) +
      # change the order of age groups in legend so they appear youngest to oldest
      scale_color_discrete(breaks=age_descriptions) +
      theme_minimal()
```

```{r}
# bar chart of age with all geographies in the most recent year
age %>%
  # only keep most recent year
  filter(year == max(.$year)) %>%
  ggplot(aes(geo_description, estimate, fill=age)) +
    geom_bar(position = "dodge", stat='identity') +
    labs(title = paste0('Poverty rate by age', ' in ', max(age$year)),
         fill = 'Age Group',
         y = 'Poverty rate (%)',
         x = '') +
    theme_minimal()
```

#### Interpretation

* The 2017 differences in poverty rates between Forsyth County and its comparison geographies are not statistically significant for any age group. 
* Within Forsyth County, children under 5 are more likely to reside in homes living in poverty than any other age group. But, the difference between children under 5 and other age groups is only statistically significant for 35 to 64 year olds and 65 and older.

#### Data

Age data is below.

```{r include = to_include}
age_data <- age %>%
  select(geo_description, year, age, estimate, moe, se, cv)

#  add to dataset for Tableau
tableau_data <- age_data %>%
    # add type of column
    mutate(type = 'Age Group',
           subtype = age) %>%
    # delete ethnicity column since this is now in the subtype column
    select(-age) %>%
    bind_rows(tableau_data, .)
  
ff_data_dt(age_data,
           c('Geographic area', 'Year', 'Age group', 'Poverty Rate', '95% MOE', 'St. Error', 'CV'))
```

#### Z-scores

The table below highlights z-scores for each age group in the past two years.

```{r include = to_include}
# enter comparison years here
age_years <- c(max(age$year), max(age$year)-1)

age %>%
  filter(year %in% age_years) %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'geo_description', 'age'))
```

Here are the age comparisons for all years within Forsyth County.

```{r include = to_include}
age %>%
  filter(geo_description == 'Forsyth') %>%
  ff_acs_zscore_dt('estimate', 'moe', c('year', 'age'))
```

### Tableau dataset

This is the final dataset that can be imported into Tableau. Users can copy it into an excel file or download a csv file.

```{r}
# estimates are percentages displayed in tens (60.45)
# convert to numbers to decimal percentages (.6045)
tableau_data %>%
  mutate_at(vars(estimate:se), funs(./100)) %>%
  # reorder column and do not include unneded columns
  select(year, geo_description, type, subtype, estimate) %>%
  # create table
  ff_data_dt(c('Year', 'Geographic area', 'Type', 'Subtype', 'Estimate'), for_tableau=TRUE)
```